---
title: "L26: Extending the chi-square to two-way tables"
output: html_document
---

```{r load-libraries, include=FALSE, message=FALSE, warning=FALSE}
library(tibble)
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(fig.align = "center", message = FALSE, warning = FALSE)
options(width = 90)
```

Every calculation from L26 in one runnable place. Last lecture had **one** categorical
variable and a null that told us the share of each category. Today there are **two**
categorical variables and the null is that they are unrelated.

There is no external data file. Every number below is typed in from the slides.
Try the practice problems at the bottom before checking the answers.

## What changes and what does not

| | Goodness of fit (L25) | Test of independence (L26) |
|---|---|---|
| Variables | one | two |
| Where $E_i$ comes from | given to you as $n \times p_i$ | computed from the **margins** |
| Test statistic | $\sum (O_i-E_i)^2/E_i$ | **identical** |
| Degrees of freedom | $k - 1$ | $(r-1)(c-1)$ |

The degrees of freedom is the part people get wrong. For a 2x2 table it is
$(2-1)(2-1) = 1$, **not** $4 - 1 = 3$.

Because the difference is squared, $(E_i - O_i)^2$ and $(O_i - E_i)^2$ give exactly
the same answer. Use whichever you remember.

## Where the expected counts come from

If two variables are independent then $P(A \text{ and } B) = P(A)P(B)$. Multiply that
probability by $n$ and you have an expected count. Doing the algebra once gives the
shortcut we actually use:

$$E_i = \frac{\text{row total} \times \text{col total}}{\text{overall total}}$$

## Example 1: the gastroenteritis outbreak

263 people, each classified two ways: did they eat the sandwich, and did they get ill.

```{r outbreak-table}
two_way <- tribble(~ sandwich, ~ nosandwich,
                     109,        4,   # row for Illness
                     116,       34)   # row for no Illness
two_way
```

### Expected counts from the margins

```{r outbreak-expected}
counts <- as.matrix(two_way)
row_totals <- rowSums(counts)
col_totals <- colSums(counts)
n          <- sum(counts)

expected <- outer(row_totals, col_totals) / n
round(expected, 2)
```

Check the conditions before going further. This is a 2x2 table, so **all four** cells
need an expected count of at least 5.

```{r outbreak-conditions}
min(expected)   # smallest expected count
```

16.33, so we are fine.

Notice the expected counts preserve the margins. That is always true and is a free
way to catch a slip.

```{r outbreak-margin-check}
rowSums(expected)
colSums(expected)
```

### The statistic by hand

```{r outbreak-by-hand}
contributions <- (expected - counts)^2 / expected
round(contributions, 4)

chi_sq <- sum(contributions)
chi_sq
```

The biggest contribution comes from the cell where we expected about 16 people and
saw only 4. That small cell does more work than any other, which is exactly what
dividing by the expected count is designed to produce.

```{r outbreak-pvalue}
# df = (2-1)(2-1) = 1
pchisq(q = chi_sq, df = 1, lower.tail = FALSE)
```

### The one-line version

```{r outbreak-chisq}
# not using Yates' correction for continuity
chisq.test(two_way, correct = FALSE)
```

If you did the arithmetic by hand off the rounded slide values you got 18.9973 rather
than 19.0742. The difference is entirely rounding. R never rounds, so R is right.

## Yates' continuity correction

The chi-square is a continuous distribution but counts are discrete. In a 2x2 table
with a small sample that approximation is at its weakest, so the correction subtracts
0.5 from each absolute gap before squaring.

$$\chi^2 = \sum_{i=1}^k \frac{(|E_i - O_i| - 0.5)^2}{E_i}$$

```{r outbreak-corrected}
# using Yates' correction (this is R's default)
chisq.test(two_way, correct = TRUE)
```

The correction always makes the statistic **smaller** and the p-value **larger**. It is
deliberately conservative. A common rule of thumb is to use it when $n < 100$ or any
$O_i < 10$. Our table has an observed count of 4, so by that rule it applies here.

## The chi-square and the two-sample z test are the same test

For a 2x2 table, $\chi^2 = z^2$ exactly.

```{r chi-vs-z}
x <- c(109, 4)     # ill, in each group
m <- c(225, 38)    # group sizes: sandwich, no sandwich

p1 <- x[1]/m[1]
p2 <- x[2]/m[2]
p_pooled <- sum(x)/sum(m)

z <- (p1 - p2) / sqrt(p_pooled * (1 - p_pooled) * (1/m[1] + 1/m[2]))
z
z^2

chisq.test(two_way, correct = FALSE)$statistic
```

Not approximately equal. Identical. The z test gives you a direction and a confidence
interval; the chi-square generalises to tables larger than 2x2.

## Example 2: HPV status and age group (a 6x2 table)

Nothing about the method required the table to be small.

```{r hpv-table}
n_way <- tribble(~ HPV, ~ noHPV,
                   160, 492,
                    85, 104,
                    48, 126,
                    90, 238,
                    82, 242,
                    50, 204)
n_way
```

### The picture first

```{r hpv-plot, out.width="80%", fig.width=8, fig.height=3.6}
age.group <- rep(c("14-19", "20-24", "25-29", "30-39", "40-49", "50-59"), 2)
hpv.test  <- c(rep("positive", 6), rep("negative", 6))
number.of.women <- c(160, 85, 48, 90, 82, 50,
                     492, 104, 126, 238, 242, 204)

hpv.data <- data.frame(age.group, hpv.test, number.of.women)

hpv.data <- hpv.data %>% group_by(hpv.test) %>%
  mutate(sum.women = sum(number.of.women),
         prop.women.age = number.of.women / sum.women)

ggplot(hpv.data, aes(x = age.group, y = prop.women.age)) +
  geom_bar(aes(fill = hpv.test), stat = "identity", position = "dodge") +
  theme_minimal(base_size = 12) +
  labs(y = "Proportion", x = "Age", fill = "HPV test",
       title = "Conditional distribution of age by HPV status")
```

`stat = "identity"` tells R the bar heights are already calculated, so it should plot
the numbers as given rather than counting rows.

Look at the 20-24 band. That one age group is going to dominate the test statistic.

### Expected counts and contributions

```{r hpv-expected}
m_hpv    <- as.matrix(n_way)
expected_hpv <- outer(rowSums(m_hpv), colSums(m_hpv)) / sum(m_hpv)
round(expected_hpv, 2)

min(expected_hpv)   # smallest expected count, conditions easily met
```

```{r hpv-contributions}
contrib_hpv <- (expected_hpv - m_hpv)^2 / expected_hpv
round(contrib_hpv, 4)

sum(contrib_hpv)
```

Read that table rather than just summing it. Most age bands contribute almost nothing.
The 20-24 row contributes about 31.8 of the 40.55 all by itself, exactly as the plot
suggested.

```{r hpv-pvalue}
# df = (6-1)(2-1) = 5
pchisq(40.55353, df = 5, lower.tail = FALSE)
```

### The one-line version

```{r hpv-chisq}
chisq.test(n_way, correct = FALSE)
```

The same function that handled a 2x2 table handles a 6x2 table with no modification.
R works out the dimensions, all twelve expected counts, and the degrees of freedom
by itself.

---

# Practice problems

### Problem 1

A study of 200 people records whether they exercise regularly and whether they have
high blood pressure:

| | High BP | Normal BP |
|---|---|---|
| Exercises | 20 | 80 |
| Does not exercise | 40 | 60 |

Compute the expected counts, check the conditions, and test for independence.

*Hint: use the row total times column total over the grand total.*

```{r q1}
tbl1 <- matrix(c(20, 80, 40, 60), nrow = 2, byrow = TRUE)
e1 <- outer(rowSums(tbl1), colSums(tbl1)) / sum(tbl1)
e1
min(e1)

chisq.test(tbl1, correct = FALSE)
```

**Answer.** Expected counts are 30, 70, 30 and 70. All are above 5, and this is a 2x2
table so that is the requirement. $\chi^2 = 9.52$ on 1 degree of freedom, p-value
about 0.002. Strong evidence that exercise and blood pressure are not independent.

### Problem 2

For Problem 1, what are the degrees of freedom and why?

**Answer.** $(2-1)(2-1) = 1$. Once the margins are fixed, filling in a single inner
cell forces the other three by subtraction, so only one cell is genuinely free to vary.
It is **not** $4-1 = 3$; that is the goodness of fit rule from L25.

### Problem 3

Rerun Problem 1 with Yates' correction. Does the conclusion change? Should you have
used it?

```{r q3}
chisq.test(tbl1, correct = TRUE)
```

**Answer.** The statistic falls to 8.60 and the p-value rises to about 0.0034, as the
correction always does. The conclusion is unchanged. By the rule of thumb the
correction is not needed here, since $n = 200$ and every observed count is well above
10.

### Problem 4

Verify for Problem 1 that $\chi^2 = z^2$.

```{r q4}
x4 <- c(20, 40)      # high BP in each group
m4 <- c(100, 100)    # group sizes
p_pool4 <- sum(x4)/sum(m4)
z4 <- (x4[1]/m4[1] - x4[2]/m4[2]) /
       sqrt(p_pool4 * (1 - p_pool4) * (1/m4[1] + 1/m4[2]))
z4
z4^2
```

**Answer.** $z = -3.086$ and $z^2 = 9.52$, matching the uncorrected chi-square exactly.
The sign of $z$ depends only on which group you typed first, which is why the
chi-square, being squared, cannot tell you direction.

### Problem 5

A 3x4 table has how many degrees of freedom?

**Answer.** $(3-1)(4-1) = 6$.
