---
title: "Code from summer session lecture 2"

---

Note that there is a good ammount of code in this lecture that you do not need to know for exams/quizzes - we are including it here in case you are interested, but you do not need to understand or be able to use all of the code in this lecture.  In this chunk I am loading all of the packages you need for the code in the markdown.  It's a usefull habit to get into to load all of the needed pacakges at the beginning of your markdown file. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(forcats)
library(fGarch) # don't need to know what this library does (used to make the skewed distributions)
```

## Visualizations for categorical data


### Example: infectious disease data

- Task: Make a bar chart of the percent of cases on infectious disease for each 
category of disease.
- First, read and view the infectious disease data from Baldi and Moore:

```{r make-bar-chart}
id_data <- read_csv("Ch01_ID-data.csv")
```

### Example: infectious disease data
```{r show-id}
id_data
```

### Example: infectious disease data
- Note the variables `number_cases` and `percent_cases`
- What do you want the bar chart to display? What is the x and y variables for 
a bar chart?


## Introducing ggplot

### First step to building a `ggplot()`: set up the canvas

- The first line of code below pulls in the ggplot package
- The second line of code below specifies the `data` set and what goes on the `x` and `y` axes

```{r, fig.height = 4, fig.width = 9.5, echo=FALSE}
library(ggplot2)
ggplot(id_data, aes(x = disease, y = percent_cases)) 
```

### Next choose a function

- We will use a `geom_` function to create our chart


`ggplot()`'s `geom_bar()` makes a bar chart


### Syntax for bar charts 

ggplot(id_data, aes(x = disease, y = percent_cases)) + 

  geom_bar(stat = "identity") 
  
 stat = "identity" tells geom_bar that we supplied a y variable that is exactly what we want to plot. 
 
 We do not need geom_bar() to calculate the number or percent for us.
 
### `ggplot()`'s `geom_bar()` makes a bar chart

```{r, fig.height = 4, fig.width = 9.5, echo=FALSE}
ggplot(id_data, aes(x = disease, y = percent_cases)) + 
  geom_bar(stat = "identity")
```

### some additions to ggplot for style


```{r, fig.height = 4, fig.width = 9.5, echo=FALSE}
ggplot(id_data, aes(x = disease, y = percent_cases)) + 
  geom_bar(stat = "identity") +
   labs(y = "Percent", x = "") +
  theme_minimal(base_size = 15) 
```

 base_size controls the font size on these plots
 
 theme_minimal affects the "look" of the plot it removes the grey background and adds grey gridlines



### `fct_reorder` reorders `disease` according to value of `percent_cases`

```{r bar-ordered, fig.height = 4, fig.width = 9.5}
library(forcats)  ##notice that we have to pull in the package forcats here!
id_data <- id_data %>% 
  mutate(disease_ordered = fct_reorder(disease, percent_cases, .desc = T))
```

### Re-ordered plot
```{r ordered, fig.height = 4, fig.width = 9.5, echo=FALSE}
ggplot(id_data, aes(x = disease_ordered, y = percent_cases)) + 
  geom_bar(stat = "identity") + 
  labs(y = "Percent", x = "") +
  theme_minimal(base_size = 15)
```

### Use `aes(fill = type)` to link the bar's fill to the disease type

  geom_bar(stat = "identity", aes(fill = type)) +

   theme(legend.position = "top")

### Use `aes(fill = type)` to link the bar's fill to the disease type
```{r bar-coloured, echo=FALSE, fig.height = 5, fig.width = 9.5}
ggplot(id_data, aes(x = disease_ordered, y = percent_cases)) + 
  geom_bar(stat = "identity", aes(fill = type)) +
  labs(y = "Percent", x = "") +
  theme_minimal(base_size = 15) +
  theme(legend.position = "top")
```

## Visualizing quantitative variables

### Example:  opioid state prescription rates

- The textbook gives an example using data from 2012.
- In the data folder, there is updated data from 2018. It came from the paper:
“Opioid Prescribing Rates by Congressional Districts, United States, 2016”, by 
Rolheiser et al. [link](https://ajph.aphapublications.org/doi/10.2105/AJPH.2018.304532)

### Example:  opioid state prescription rates
Problem:  To determine the extent to which opioid prescribing rates vary across US congressional districts.

Plan: In an observational cross-sectional framework using secondary data, they constructed 2016 congressional district–level opioid prescribing rate estimates using a population-weighted methodology.

Data:  In the data structure we have State as the unit of analysis, and measured perscription rates as the variable of interest


### Example:  opioid state prescription rates

```{r read-opi-data}
opi_data <- read.csv("Ch01_opioid-data.csv")
head(opi_data)
```

- `Mean` provides the mean prescribing rate per 100 individuals. Thus, a mean
of 121.31 implies that in Alabama, there were 121.31 opioid prescriptions per 100 
persons, an average across the 7 congressional districts.

### Histogram of opioid prescription rates

- Task: Make a histogram of the average prescribing rates across US states
- What is the x variable? What is the y variable? 
- What `geom` should be used?

### Histogram of opioid prescription rates - default is 30 bins
ggplot(data = opi_data, aes(x = Mean)) + 

  geom_histogram(col = "white") +

  labs(x = "Mean opioid prescription rate (per 100 individuals)", 

       y = "Number of states") +

  theme_minimal(base_size = 15)
  
### Histogram of opioid prescription rates
```{r hist-binwidth-no-width-specified, fig.height = 4, fig.width = 8, echo=FALSE}
ggplot(data = opi_data, aes(x = Mean)) + 
  geom_histogram(col = "white") +
  labs(x = "Mean opioid prescription rate (per 100 individuals)", 
       y = "Number of states") +
  theme_minimal(base_size = 15)
```


### same graph, change the bins `geom_histogram(binwidth = 5)`
```{r hist-binwidth-5, fig.height = 4, fig.width = 8, echo=FALSE}
ggplot(data = opi_data, aes(x = Mean)) + 
  geom_histogram(col = "white", binwidth = 5) +
  labs(x = "Mean opioid prescription rate (per 100 individuals)", 
       y = "Number of states") +
  theme_minimal(base_size = 15)
```


### change the bins again `geom_histogram(binwidth = 10)`
```{r hist-binwidth-10, fig.height = 4, fig.width = 8, echo=FALSE}
ggplot(data = opi_data, aes(x = Mean)) + 
  geom_histogram(col = "white", binwidth = 10) +
  labs(x = "Mean opioid prescription rate (per 100 individuals)", 
       y = "Number of states") +
  theme_minimal(base_size = 15)
```


## Statistics is Everywhere


### Skewed data

```{r, echo = F, out.width="80%", out.height="80%", fig.align='center'}
# STUDENTS: YOU DON'T NEED TO KNOW THE CODE IN THIS CODE CHUNK USED TO MAKE THIS PLOT.
x <- seq(0,1,length=100)
db <- dbeta(x, 2, 5)
db2 <- dbeta(x, 8, 2)

ggplot() + 
  geom_line(aes(x,db, col = "skewed right"), lwd = 2) + 
  geom_line(aes(x,db2, col = "skewed left"), lwd = 2) +
  labs(y = "Density") + theme_minimal(base_size = 18) + 
  theme(legend.title = element_blank())
```    

### Apartment rent in SF

Data:  Here I will present data that I simulated in r using a mean value published on rentjungle.com
- you will not be expected to do this or be tested on it.


```{r distribution-ex, echo = F, message=FALSE, warning=FALSE}
# This code chunk simulates rent price distributions under varying assumptions.
# Note to students: you won't be tested on your understanding of this code chunk.

set.seed(1236)
rent_n <- 1000
rent_mean <- 3300
rent_sd <- 500
rent_skew_factor <- 5
rent_sym <- rnorm(n = rent_n, mean = rent_mean, sd = rent_sd)
rent_sym_out_left <- c(rent_sym[ seq(1, rent_n - 3) ], 1200, 1250, 1500)
rent_sym_out_right <- rent_sym + c(rep(0, rent_n - 3), 6000, 5000, 5500)
rent_left_skew <- rsnorm(
  n = rent_n, mean = rent_mean,
  sd = rent_sd, xi = 1 / rent_skew_factor
)
rent_right_skew <- rsnorm(
  n = rent_n, mean = rent_mean,
  sd = rent_sd, xi = rent_skew_factor
)
rent_bimodal <- c(
  rnorm(n = rent_n / 2, mean = rent_mean, sd = 300),
  rnorm(n = rent_n / 2, mean = 2000, sd = 300)
)

rent_data <- data.frame(
  sym = rent_sym,
  left_skew = rent_left_skew,
  sym_out_left = rent_sym_out_left,
  sym_out_right = rent_sym_out_right,
  right_skew = rent_right_skew,
  bimodal = rent_bimodal
)

rm(rent_sym, rent_left_skew, rent_sym_out_left, 
   rent_sym_out_right, rent_right_skew, rent_bimodal)

rent_summs <- rent_data %>%
  summarize_all(list(mean = mean, median=median))

### now sample only 100 apartments
rent_smaller_n <- 100

rent_sym_small <- rnorm(n = rent_smaller_n, mean = rent_mean, sd = rent_sd)
rent_sym_out_left_small <- c(rent_sym_small[ seq(1, rent_smaller_n - 3) ], 1200, 1250, 1500)
rent_sym_out_right_small <- rent_sym_small + c(rep(0, rent_smaller_n - 3), 6000, 5000, 5500)

rent_data_small <- data.frame(
  sym = rent_sym_small,
  sym_out_left = rent_sym_out_left_small,
  sym_out_right = rent_sym_out_right_small
)

rent_summs_small <- rent_data_small %>%
  summarize_all(funs(mean, median))

rm(rent_sym_out_left_small, rent_sym_out_right_small, rent_sym_small)
```

### Example: Apartment rent in SF
Suppose that the distribution of rent prices looked like this. The green ticks underneath the histograms shows you the exact rent values that 
contribute data to each bin. 
STUDENTS you do NOT need to be able to understand the code in the below chunk

```{r sym-plot, echo = F, fig.width=7, fig.height=4}
sym_plot <- ggplot(rent_data, aes(x = sym)) +
  geom_histogram(binwidth = 200, col = "white") +
  labs(x = "Monthly Rent ($)", 
       y = "Count of apartments",
       title = "Symmetric distribution in rental prices ($)") +
  geom_vline(xintercept = rent_summs$sym_mean) +
  geom_vline(xintercept = rent_summs$sym_median, col = "blue") +
  geom_text(aes(x = rent_summs$sym_mean + 150, y = 20),
            label = "Mean", check_overlap = T) +
  geom_text(aes(x = rent_summs$sym_median - 150, y = 20),
            label = "Median", check_overlap = T, col = "blue") +
  geom_rug(col = "#005824", alpha = 0.5) + #hex colours (see link below)
  theme_minimal(base_size = 15)

sym_plot
```

[Hex color picker](https://www.google.com/search?q=hex+colors+picker&oq=hex+colour&aqs=chrome.2.69i57j0l5.3494j0j1&sourceid=chrome&ie=UTF-8)

### Example: Apartment rent in SF
Summarizing numerically:
Center: 

```{r numerics}
# in base R
mean(rent_data[,"sym"])
median(rent_data[,"sym"])
```

```{r}
# using the summarize function and a pipe operator 
rent_data %>% summarize(
  mean=mean(sym),
  median = median(sym))
```

## Mean vs Median:  Outliers and sample size, skew, shape 

### Example: Apartment rent in SF
Now suppose that there were three rents within the data set with much larger 
values than the rest of the distribution. Here is the plot for this updated data.

```{r sym-outlier-high-plot, echo = F, fig.width=9, fig.height=4}
sym_out_right_plot <- ggplot(rent_data, aes(x = sym_out_right)) +
  geom_histogram(binwidth = 200, col = "white") +
  xlab("Monthly Rent ($)") + ylab("Count") +
  ggtitle("Symmetric, but with outliers on the right, n=1000") +
  geom_vline(xintercept = rent_summs$sym_out_right_mean) +
  geom_vline(xintercept = rent_summs$sym_out_right_median, col = "blue") +
  geom_text(aes(x = rent_summs$sym_out_right_mean + 350, y = 20),
            label = "Mean", check_overlap = T) +
  geom_text(aes(x = rent_summs$sym_out_right_median - 350, y = 20),
            label = "Median", check_overlap = T, col = "blue") +
  geom_rug(col = "#005824", alpha = 0.5) +
  theme_minimal(base_size = 15)

sym_out_right_plot
```

- With 1000 sampled points the outliers do not have a large effect on the mean

### Example: Apartment rent in SF
Imagine instead, there were only 100 sampled points. Here, the outliers have a 
larger effect on the mean. **The mean is not resistant to outliers.**

```{r sym-outlier-high-smaller-n, echo = F, fig.width=9, fig.height=4}
sym_out_right_plot <- ggplot(rent_data_small, aes(x = sym_out_right)) +
  geom_histogram(binwidth = 400, col = "white") +
  xlab("Monthly Rent ($)") + ylab("Count") +
  ggtitle("Symmetric, but with outliers on the right, n=100") +
  geom_vline(xintercept = rent_summs_small$sym_out_right_mean) +
  geom_vline(xintercept = rent_summs_small$sym_out_right_median, col = "blue") +
  geom_text(aes(x = rent_summs_small$sym_out_right_mean + 350, y = 20),
            label = "Mean", check_overlap = T) +
  geom_text(aes(x = rent_summs_small$sym_out_right_median - 350, y = 20),
            label = "Median", check_overlap = T, col = "blue") +
  geom_rug(col = "#005824", alpha = 0.5) +
  theme_minimal(base_size = 15)

sym_out_right_plot
```

### Example: Apartment rent in SF
Consider instead what happens if there are many high-end apartments in the area. Here is the histogram of data for this example:

```{r right-skew-plot, echo = F, fig.width=9, fig.height=4}
right_skew_plot <- ggplot(rent_data, aes(x = right_skew)) +
  geom_histogram(binwidth = 150, col = "white") +
  xlab("Monthly Rent ($)") + ylab("Count") +
  ggtitle("Right-Skewed") +
  geom_vline(xintercept = rent_summs$right_skew_mean) +
  geom_vline(xintercept = rent_summs$right_skew_median, col = "blue") +
  geom_text(aes(x = rent_summs$right_skew_mean + 150, y = 20), label = "Mean", check_overlap = T) +
  geom_text(aes(x = rent_summs$right_skew_median - 150, y = 20), label = "Median", check_overlap = T, col = "blue") +
  geom_rug(col = "#005824", alpha = 0.5) +
  theme_minimal(base_size = 15)
right_skew_plot
```


### Example: Apartment rent in SF
Now, suppose that the sample of estimates did not look like the distribution in the previous example. Instead, it looked like this: 

```{r bimodal-plot, echo = F, fig.width=9, fig.height=4}
bi_plot <- ggplot(rent_data, aes(x = bimodal)) +
  geom_histogram(binwidth = 250, col = "white") +
  xlab("Monthly Rent ($)") + ylab("Count") +
  ggtitle("Bimodal") +
  geom_vline(xintercept = rent_summs$bimodal_mean) +
  geom_vline(xintercept = rent_summs$bimodal_median, col = "blue") +
  geom_text(aes(x = rent_summs$bimodal_mean + 150, y = 20),
            label = "Mean", check_overlap = T) +
  geom_text(aes(x = rent_summs$bimodal_median - 150, y = 20),
            label = "Median", check_overlap = T, col = "blue") +
  geom_rug(col = "#005824", alpha = 0.5) +
  theme_minimal(base_size = 15)
bi_plot
```

## Measures of spread

### Quantiles using R
`quantile(variable, 0.25)`

```{r quantiles}
rent_data %>% summarize(
  Q1 = quantile(sym, 0.25),
  median = median(sym),
  Q3 = quantile(sym, 0.75)
  )
```

### R's quantile function:  Note

- `quantile(variable, 0.25)` will not always give the exact same answer you 
calculate by hand
- The R function is optimized for its statistical properties and is slightly 
different than the book's method
- To get the exact same answer as by hand use `quantile(data, 0.25, type = 2)`
- You may use either one in this class. Most commonly, people do not specify
`type=2`

### `dplyr`'s summarize() to calculate the five number summary
Using our original example of rent data:

```{r five-number-summary2}
rent_data %>% summarize(
  min = min(sym),
  Q1 = quantile(sym, 0.25),
  median = median(sym),
  Q3 = quantile(sym, 0.75),
  max = max(sym)
)
```

## Example:  Hospital cesarean delivery rates

### Example:  Hospital cesarean delivery rates
These data were provided by the first author (Kozhimannil) of a manuscript 
published in the journal *Health Affairs*. [link](https://www.healthaffairs.org/doi/10.1377/hlthaff.2012.1030)

### Example:  Hospital cesarean delivery rates
We start by importing the data:

```{r import-tidy-cs-data, message=FALSE, warning=FALSE}
library(readxl) 
# this library helps with reading xlsx and xls files into R
CS_dat <- read_xlsx("Kozhimannil_Ex_Cesarean.xlsx", sheet = 1)
```

### Example:  Hospital cesarean delivery rates

```{r import-tidy-cs-data2}
head(CS_dat)
```

### Example:  Hospital cesarean delivery rates

```{r import-tidy-cs-data3}
names(CS_dat) 
```

let's take a moment to discuss variable names containing spaces

### Sidenote on variable names containing spaces

- Two variables in `CS_dat` contain spaces. 
- We generally want to remove spaces from variable names. 

* Question: Which `dplyr` function can we use to change the variable names?
* Answer: `rename(new_name = old_name)` can be used. When the old variable name contains spaces, you need to place back ticks around it like this: 


```{r rename-vars-with-spaces}
CS_dat <- CS_dat %>% rename(cs_rate = `Cesarean rate *100`,
                            low_risk_cs_rate = `Low Risk Cearean rate*100`)
```

- See [this paper](https://www.tandfonline.com/doi/abs/10.1080/00031305.2017.1375989) for tips on storing data in Excel for later analysis. 

### Tidy the data for analysis

For our example, we are only interested in each hospital's cesarean delivery rate,
the rate for lower risk pregnancies, and the number of births at the hospital.

```{r tidy-the-data}
CS_dat <- CS_dat %>%
  select(Births, cs_rate,low_risk_cs_rate) %>%
  rename(num_births = Births)
```

### Histogram of cesarean delivery rates across US hospitals
```{r histogram-cs, fig.width=8, fig.height=4,echo=F}
ggplot(CS_dat, aes(x = cs_rate)) +
  geom_histogram(col = "white", binwidth = 5) +
  labs( x = "Cesarean delivery rate (%)", y = "Count",
    caption = "Data from: Kozhimannil, Law, and Virnig. Health Affairs. 2013;32(3):527-35.") +
  geom_rug(alpha = 0.2, col = "forest green") + #alpha controls transparency
  theme_minimal(base_size = 15)
```

### Spread of cesarean delivery rates across US hospitals
- What can you say about this distribution? Would you expect so much variation 
across hospitals in their rates of cesarean delivery? 
- Let's describe the **spread** of these data using the methods from Chapter 2.


### Quantiles 

```{r quantiles2}
CS_dat %>% summarize(
  Q1 = quantile(cs_rate, 0.25),
  median = median(cs_rate),
  Q3 = quantile(cs_rate, 0.75)
  )
```


### `dplyr`'s summarize() to calculate the five number summary
```{r five-number-summary}
CS_dat %>% summarize(
  min = min(cs_rate),
  Q1 = quantile(cs_rate, 0.25),
  median = median(cs_rate),
  Q3 = quantile(cs_rate, 0.75),
  max = max(cs_rate)
)
```

### Histogram of low risk cesarean delivery rates across US hospitals
```{r histogram-lcs, fig.width=8, fig.height=4,echo=F}
ggplot(CS_dat, aes(x = low_risk_cs_rate)) +
  geom_histogram(col = "white", binwidth = 5) +
  labs( x = "Low risk Cesarean delivery rate (%)", y = "Count",
    caption = "Data from: Kozhimannil, Law, and Virnig. Health Affairs. 2013;32(3):527-35.") +
  geom_rug(alpha = 0.2, col = "forest green") + #alpha controls transparency
  theme_minimal(base_size = 15)
```

### `dplyr`'s summarize() to calculate the five number summary
```{r five-number-summaryL}
CS_dat %>% summarize(
  min = min(low_risk_cs_rate),
  Q1 = quantile(low_risk_cs_rate, 0.25),
  median = median(low_risk_cs_rate),
  Q3 = quantile(low_risk_cs_rate, 0.75),
  max = max(low_risk_cs_rate)
)
```


### `dplyr`'s summarize() to calculate the standard deviation and the variance
```{r five-number-summary-plus, echo = T}
CS_dat %>% summarize(
  cs_sd = sd(cs_rate),
  cs_var = var(cs_rate)
)
```



### Box plots in R 


### Box plots provide a nice visual summary of the center and spread
```{r box-plot-ex, width = 2, fig.width = 4.5, fig.height = 5.5, echo=F}
ggplot(CS_dat, aes(y = cs_rate)) +
  geom_boxplot() +
  ylab("Cesarean delivery rate (%)") +
  labs(title = "Box plot of the CS rates across \nUS hospitals",
       caption = "Data from: Kozhimannil et al. 2013.") +
  theme_minimal(base_size = 15) +
  scale_x_continuous(labels = NULL) 
  # removes the labels from the x axis
```


### We only skimmed the surface!

- Here is some extra material for those of you who love data visualization. This material won't be tested.
    - [RStudio ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
    - [Kieran Healy's data visualization book](https://www.amazon.com/gp/product/0691181624/ref=as_li_tl?ie=UTF8&tag=kieranhealysw-20&camp=1789&creative=9325&linkCode=as2&creativeASIN=0691181624&linkId=16d53b3cc1ec3bc3aac60b27c29b92e8)

