---
title: "L27: Wrapping up and structured review"
output: html_document
---

```{r load-libraries, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(fig.align = "center", message = FALSE, warning = FALSE)
options(width = 90)
```

Every calculation from L27 in one runnable place, plus the decision tree written out as a
lookup table you can revise from.

There is no external data file. Every number below is typed in from the slides.

The exam does not tell you which test to use. That is the whole skill this lecture is
about, so the practice problems at the bottom are mostly **choose the test** problems
rather than arithmetic. Try them before checking the answers.

## The two recipes

Everything in Part III is one of these two, with different ingredients.

**Recipe 1, hypothesis testing.** Check assumptions, build a ratio of effect to
variability, convert that ratio to a probability, decide.

**Recipe 2, confidence intervals.** Take the estimate, find the critical value, multiply it
by the variability, add and subtract.

Every test statistic in the course has the same shape:

$$\text{test statistic} = \frac{\text{effect}}{\text{variability}} = \frac{\text{signal}}{\text{noise}}$$

## The decision tree as a lookup table

Four questions pick the test. What kind of outcome, how many groups, paired or
independent, parametric or not.

### Continuous outcomes

| Situation | Test | R |
|---|---|---|
| 1 group vs a hypothesized value | one sample $t$ | `t.test(x, mu = )` |
| 1 group, $\sigma$ known | one sample $z$ | `pnorm()` |
| 2 groups, independent | two sample $t$ | `t.test(y ~ g)` |
| 2 groups, independent, assumptions fail | Wilcoxon rank sum | `wilcox.test(y ~ g)` |
| 2 groups, paired | paired $t$ | `t.test(x1, x2, paired = TRUE)` |
| 2 groups, paired, assumptions fail | Wilcoxon signed rank | `wilcox.test(x1, x2, paired = TRUE)` |
| 3+ groups | ANOVA, then Tukey or Bonferroni | `aov()`, `TukeyHSD()` |
| 3+ groups, assumptions fail | Kruskal-Wallis | `kruskal.test(y ~ g)` |
| continuous predictor | correlation, linear regression | `cor.test()`, `lm()` |

### Categorical outcomes

| Situation | Test | R |
|---|---|---|
| 1 group, 2 categories | one sample $z$ for a proportion | `prop.test(x, n, p = )` |
| 1 group, 3+ categories | chi-square goodness of fit | `chisq.test(counts, p = )` |
| 2 groups, 2 categories | two sample $z$, or chi-square | `prop.test(x, n)`, `chisq.test()` |
| 2 categorical variables, any table | chi-square test of independence | `chisq.test(table)` |

The paired versus independent question is the one people miss. If the **same subject**
appears in both conditions, or subjects are matched in pairs, the data are paired.

## Example 1: staph infections

917 patients randomised to a staph-killing nasal ointment or placebo. 17 of 504 in the
treatment group developed a post-surgical infection, 32 of 413 in the placebo group.

Categorical outcome, two independent groups, so this is a two sample $z$ test for
proportions. The question asks about a *reduction*, so the test is one sided.

```{r staph-setup}
x <- c(17, 32)      # infections
n <- c(504, 413)    # group sizes

p1 <- x[1] / n[1]
p2 <- x[2] / n[2]
c(p1 = p1, p2 = p2, difference = p1 - p2)
```

### Ingredient 2, and why we pool

The null says $p_1 = p_2$. If that is true, splitting the data in two to estimate that
common proportion throws information away, so we pool.

$$\hat{p} = \frac{17+32}{504+413}, \qquad SE = \sqrt{\hat{p}(1-\hat{p})\Big(\frac{1}{n_1}+\frac{1}{n_2}\Big)}$$

```{r staph-pooled}
p_pooled <- sum(x) / sum(n)
p_pooled

se_pooled <- sqrt(p_pooled * (1 - p_pooled) * (1/n[1] + 1/n[2]))
se_pooled
```

### The test statistic and the p-value

```{r staph-z}
z <- (p1 - p2) / se_pooled
z

# one sided, and pnorm gives the left tail by default
pnorm(z)
```

### The one-line version

```{r staph-prop-test}
prop.test(x = x, n = n, alternative = "less", correct = FALSE)
```

Two things to notice in that output. The p-value matches the hand calculation. And
`X-squared` is 8.5906, whose square root is our $z$ of 2.931. On a 2x2 table
$\chi^2 = z^2$ exactly, which is why R can report either one.

```{r staph-chi-vs-z}
z^2
```

### The confidence interval, and why we do *not* pool

This is the single most common mistake on this material.

A confidence interval is not computed under the null, so the assumption that let us pool is
gone. Each group gets its own proportion.

$$(\hat{p}_1 - \hat{p}_2) \pm z^*\sqrt{\frac{\hat{p}_1(1-\hat{p}_1)}{n_1} + \frac{\hat{p}_2(1-\hat{p}_2)}{n_2}}$$

```{r staph-ci}
se_unpooled <- sqrt(p1 * (1 - p1) / n[1] + p2 * (1 - p2) / n[2])
se_unpooled

(p1 - p2) + c(-1, 1) * 1.96 * se_unpooled
```

So the 95% CI is about -0.0740 to -0.0135. It excludes 0, which had to happen given the
small p-value.

Here is what pooling by mistake would have given you:

```{r staph-ci-wrong}
(p1 - p2) + c(-1, 1) * 1.96 * se_pooled   # NOT the right interval
```

Close, but wrong, and on an exam it is wrong for a reason you are expected to be able to
state. Use the pooled SE for the **test**, the unpooled SE for the **interval**.

Read the interval in plain language. The data are consistent with anything from about 1.4
to about 7.4 fewer infections per hundred patients.

## Example 5: births by day of the week

700 births, classified by day. One sample, one categorical variable with seven categories,
so this is a chi-square **goodness of fit** test. The null is that all seven days are
equally likely.

```{r births}
births <- c(110, 124, 104, 94, 112, 72, 84)
names(births) <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")
births
sum(births)
```

```{r births-expected}
expected <- sum(births) / 7   # 100 per day under the null
expected

min(expected)   # condition: every expected count at least 5
```

```{r births-test}
# chisq.test assumes equal probabilities by default, which is exactly our null
chisq.test(births)
```

Degrees of freedom are $k - 1 = 6$. Not $(r-1)(c-1)$, because there is only one variable
here.

Look at where the statistic comes from rather than just reading the total:

```{r births-contributions}
contrib <- (births - expected)^2 / expected
round(contrib, 3)
```

Saturday alone contributes 7.84 of the 19.12. The test says births are not uniform across
the week, but it cannot tell you whether that is biology or scheduled deliveries.

## Example 6: grades

350 Berkeley transcripts, compared against a California-wide distribution of A = 50%,
B = 30%, C = 15%, F = 5%.

One sample, one categorical variable, and the null comes from an outside source. Goodness
of fit again, **not** a test of independence, because there is only one variable.

```{r grades}
observed <- c(A = 224, B = 99, C = 17, F = 10)
p_null   <- c(0.50, 0.30, 0.15, 0.05)

expected_g <- sum(observed) * p_null
expected_g
```

Expected counts do not have to be whole numbers, but they must sum to the observed total.
That is a free check on your arithmetic:

```{r grades-check}
sum(expected_g) == sum(observed)
```

```{r grades-contributions}
contrib_g <- (observed - expected_g)^2 / expected_g
round(contrib_g, 3)
sum(contrib_g)
```

```{r grades-test}
chisq.test(observed, p = p_null)
```

The C category contributes more than half the statistic all by itself, because we expected
52.5 students and saw 17. Dividing by the expected count is what makes a shortfall in a
small category outweigh a surplus in a large one.

And read the direction. Berkeley has *more* As and *fewer* Cs than the comparison
distribution, which is the opposite of the rumour we set out to test.

---

# Practice problems

## Choose the test

For each scenario, answer the four questions and name the test.

### Problem 1

A researcher measures resting heart rate in 40 runners and 40 non-runners and wants to know
whether mean heart rate differs.

**Answer.** Continuous outcome, two groups, independent (different people), assumptions
plausible at n = 40 per group. **Two sample $t$ test.** `t.test(hr ~ group)`

### Problem 2

The same 30 patients have their blood pressure measured before and after eight weeks on a
drug.

**Answer.** Continuous outcome, two measurements, and the *same* patients appear in both,
so the data are **paired**. **Paired $t$ test**, which is a one sample $t$ on the 30
differences. `t.test(before, after, paired = TRUE)`

If the differences look badly skewed at n = 30, the Wilcoxon signed rank is the
non-parametric partner.

### Problem 3

A survey of 400 people records political party (4 categories) and whether they support a
policy (yes or no). Are the two related?

**Answer.** **Two** categorical variables measured on one sample, so this is a chi-square
**test of independence** on a 4x2 table, with $(4-1)(2-1) = 3$ degrees of freedom.
`chisq.test(table(party, support))`

### Problem 4

You roll a die 600 times and want to know whether it is fair.

**Answer.** **One** categorical variable with six categories, one sample, and the null gives
the expected share of each. **Chi-square goodness of fit**, $k - 1 = 5$ degrees of freedom.

```{r q4}
rolls <- c(96, 108, 92, 118, 87, 99)
chisq.test(rolls)   # equal probabilities is the default
```

### Problem 5

Stress scores from four different treatment groups, 25 people in each, all independent.

**Answer.** Continuous outcome, four independent groups. **ANOVA**, followed by pairwise
comparisons with Tukey's HSD or a Bonferroni correction if the ANOVA is significant. ANOVA
never tells you *which* groups differ.

## Compute

### Problem 6

In a trial, 45 of 300 patients on treatment relapsed, compared with 68 of 290 on placebo.
Test whether the treatment reduces relapse, and give a 95% confidence interval for the
difference.

*Hint: pool for the test, do not pool for the interval.*

```{r q6}
x6 <- c(45, 68)
n6 <- c(300, 290)

p1_6 <- x6[1]/n6[1]
p2_6 <- x6[2]/n6[2]

# test: pooled
p_pool6  <- sum(x6)/sum(n6)
se_pool6 <- sqrt(p_pool6 * (1 - p_pool6) * (1/n6[1] + 1/n6[2]))
z6 <- (p1_6 - p2_6) / se_pool6
z6
pnorm(z6)

prop.test(x = x6, n = n6, alternative = "less", correct = FALSE)

# interval: unpooled
se_un6 <- sqrt(p1_6*(1-p1_6)/n6[1] + p2_6*(1-p2_6)/n6[2])
(p1_6 - p2_6) + c(-1, 1) * 1.96 * se_un6
```

**Answer.** $\hat{p}_1 = 0.150$ and $\hat{p}_2 = 0.2345$, a difference of -0.0845. The pooled
SE is 0.0324, giving $z = -2.61$ and a one sided p-value of 0.0046. Reject the null. The
95% CI using the unpooled SE is about -0.148 to -0.021, which excludes 0 as it must.

### Problem 7

Verify on Problem 6 that $\chi^2 = z^2$.

```{r q7}
z6^2
prop.test(x = x6, n = n6, correct = FALSE)$statistic
```

**Answer.** Both are 6.797. They are the same test. The $z$ version gives you a direction and
an interval, the chi-square version generalises to bigger tables.

### Problem 8

A clinic expects its patients to be 40% under 30, 35% aged 30 to 50, and 25% over 50. Among
200 patients it sees 60, 82 and 58. Is the clinic's mix different from what was expected?

```{r q8}
obs8 <- c(60, 82, 58)
p8   <- c(0.40, 0.35, 0.25)

sum(obs8) * p8          # expected counts
chisq.test(obs8, p = p8)
```

**Answer.** Expected counts are 80, 70 and 50, all well above 5. $\chi^2 = 8.34$ on 2
degrees of freedom, p-value about 0.016. Reject at $\alpha = 0.05$. The under-30 group is
the biggest contributor, with 60 observed against 80 expected.

Note the degrees of freedom is $k - 1 = 2$, not 3. One variable, three categories.

### Problem 9

You run a test and get p = 0.03, but your 95% confidence interval contains the null value.
What went wrong?

**Answer.** You made an arithmetic error somewhere. A two sided test at $\alpha = 0.05$ and
a 95% confidence interval are the same arithmetic run in two directions, so they can never
disagree. The usual culprits are using the wrong critical value, or pooling in the interval
when you should not have.

The one legitimate exception to check for is a **one sided** test, where a p-value below
0.05 is perfectly compatible with a two sided interval that contains the null.

### Problem 10

Which of the tests we covered are non-parametric, and what do you give up by using one?

**Answer.** The Wilcoxon rank sum (also called Mann-Whitney), the Wilcoxon signed rank, the
sign test, and Kruskal-Wallis. They work on ranks rather than raw values, so they do not
assume a normal distribution and an extreme outlier moves them much less.

What you give up is **power**. If the normality assumption really does hold, the
non-parametric version is less likely to detect a real effect than its parametric partner.
You are paying in sensitivity for robustness.
