---
title: "L20: ANOVA"
output: html_document
---

```{r load-libraries, include=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(broom)
knitr::opts_chunk$set(fig.align = "center", message = FALSE, warning = FALSE)
```

Every calculation from L20 in one runnable place. This lecture extends the
two-sample t test to **three or more** groups. Nothing here needs an external
data file, all of the data is created in the code below. Try the practice
problems at the bottom before checking the answers.

## Why not just run a lot of t tests?

With $k$ groups there are $\binom{k}{2}$ pairwise comparisons:

```{r how-many-comparisons}
choose(3, 2) # 3 groups
choose(4, 2) # 4 groups
choose(10, 2) # 10 groups
```

Each test carries its own chance of a false positive, and you end up with a pile
of p-values instead of one answer. If the null is true everywhere and you run 6
independent tests at $\alpha = 0.05$, the chance of at least one "significant"
result is:

```{r familywise}
1 - (1 - 0.05)^6
```

ANOVA replaces all of them with **one** test of **one** null hypothesis:

$$H_0: \mu_1 = \mu_2 = \dots = \mu_K \qquad
H_a: \text{at least one } \mu \text{ differs}$$

# Part 1. Seeing between-group vs. within-group variation

The whole test is a ratio:

$$F = \frac{\text{variation among group means}}{\text{variation among individuals in the same group}}$$

The four plots below vary the numerator and the denominator on purpose. Each
chunk uses `set.seed(1)`, so plot D draws the *same* standard Normal values as
plot B, only rescaled to a larger standard deviation.

A helper so we do not repeat the plotting code four times:

```{r plot-helper}
three_group_plot <- function(mu, sd) {
  set.seed(1)
  dat <- data.frame(group_1 = rnorm(10, mu[1], sd),
                    group_2 = rnorm(10, mu[2], sd),
                    group_3 = rnorm(10, mu[3], sd))

  narrow <- dat %>%
    pivot_longer(cols = group_1:group_3,
                 names_to = "Group", values_to = "Measure")

  means <- narrow %>%
    group_by(Group) %>%
    summarise(Measure = mean(Measure))

  ggplot(narrow, aes(x = Group, y = Measure)) +
    geom_point(aes(pch = "data")) +
    geom_point(data = means, col = "red", size = 2,
               aes(pch = "mean estimate")) +
    geom_text(data = means, col = "red", size = 4,
              aes(y = 40, label = paste("mean =", round(Measure, 2)))) +
    scale_shape_manual(values = c(4, 20)) +
    scale_y_continuous(limits = c(2, 42)) +
    theme_minimal(base_size = 15)
}
```

**Plot A.** All three population means are 15. Between-group variation is small,
within-group variation is not.

```{r plot-a}
three_group_plot(mu = c(15, 15, 15), sd = 3)
```

**Plot B.** Means 15, 20, 27 with the same spread. Between-group variation is now
large relative to within.

```{r plot-b}
three_group_plot(mu = c(15, 20, 27), sd = 3)
```

**Plot C.** Means 15, 20, 15. Only the middle group moves.

```{r plot-c}
three_group_plot(mu = c(15, 20, 15), sd = 3)
```

**Plot D.** The same means as plot B, but the standard deviation goes from 3 to
7. The signal did not change, the noise did, and the difference is now much
harder to see.

```{r plot-d}
three_group_plot(mu = c(15, 20, 27), sd = 7)
```

Run the ANOVA on the data behind plots B and D to see what that costs you:

```{r ab-anova}
make_data <- function(mu, sd) {
  set.seed(1)
  data.frame(group_1 = rnorm(10, mu[1], sd),
             group_2 = rnorm(10, mu[2], sd),
             group_3 = rnorm(10, mu[3], sd)) %>%
    pivot_longer(cols = group_1:group_3,
                 names_to = "Group", values_to = "Measure")
}

tidy(aov(Measure ~ Group, data = make_data(c(15, 20, 27), sd = 3)))
tidy(aov(Measure ~ Group, data = make_data(c(15, 20, 27), sd = 7)))
```

Same means, much smaller $F$, much larger p-value.

# Part 2. Descriptive plots for three or more groups

```{r descriptive-data}
diff_3_narrow <- make_data(c(15, 20, 27), sd = 7)
```

Option 1, a box plot per group with the points on top:

```{r boxplot}
ggplot(diff_3_narrow, aes(x = Group, y = Measure)) +
  geom_boxplot() +
  geom_point() +
  theme_minimal(base_size = 15)
```

Option 2, a density plot per group:

```{r density}
ggplot(diff_3_narrow, aes(x = Measure)) +
  geom_density(aes(fill = Group), alpha = 0.5) +
  theme_minimal(base_size = 15)
```

Option 3, faceted histograms:

```{r histogram}
ggplot(diff_3_narrow, aes(x = Measure)) +
  geom_histogram(aes(fill = Group), col = "white", binwidth = 2.5) +
  facet_wrap(~ Group, nrow = 3) +
  theme_minimal(base_size = 15)
```

# Part 3. The cancer example, end to end

Mice with transplanted glioma cells were randomly assigned to one of four
treatments, and tumor volume was measured after 21 days.

```{r cancer-data}
treatment <- c(rep("Irradiation", 4), rep("Cannabinoids", 5),
               rep("Both", 6), rep("Neither", 7))

tumor_volume <- c(30, 46, 46, 95,          # Irradiation
                  12, 14, 16, 41, 47,      # Cannabinoids
                  5, 4, 4, 4, 10, 9,       # Both
                  24, 30, 43, 51, 62, 32, 96) # Neither

cancer_data <- data.frame(treatment, tumor_volume)
head(cancer_data)
```

`fct_relevel()` sets the order the groups appear in on the graph:

```{r relevel}
cancer_data <- cancer_data %>%
  mutate(trt_order = fct_relevel(treatment, c("Neither", "Irradiation",
                                              "Cannabinoids", "Both")))
```

## Step 1. Summary statistics

```{r summary-stats}
summary_stats <- cancer_data %>%
  group_by(trt_order) %>%
  summarise(mean_vol = mean(tumor_volume),
            sd_vol = sd(tumor_volume),
            samp_size = n())

summary_stats
```

## Step 2. Check the conditions

1. $k$ independent SRSs. Holds by design here, the mice were randomised.
2. Each population Normal. Hard to judge with 4 to 7 mice, but ANOVA is robust
to non-Normality.
3. Equal population standard deviations. The rule of thumb is
$s_{max}/s_{min} < 2$.

```{r sd-ratio}
summary_stats %>%
  summarise(s_max = max(sd_vol), s_min = min(sd_vol),
            ratio = max(sd_vol) / min(sd_vol))
```

The ratio is about 10, so condition 3 is **violated**. The group with the
smallest mean also has the smallest spread, which is very common for a
measurement that cannot go below zero. We run the ANOVA anyway because it is the
tool this chapter teaches, but the caveat travels with the result.

## Step 3. Plot before you test

```{r cancer-plot}
set.seed(42) # geom_jitter is random, so seed it to get the same picture twice
ggplot(cancer_data, aes(x = trt_order, y = tumor_volume)) +
  geom_jitter(pch = 4, width = 0.1) +
  geom_point(data = summary_stats, aes(y = mean_vol, col = "Mean"), pch = 19) +
  labs(y = "Tumor volume (mm cubed)", x = "Treatment", col = NULL) +
  theme_minimal(base_size = 15)
```

## Step 4. Run the ANOVA

```{r cancer-anova}
cancer_anova <- aov(tumor_volume ~ treatment, data = cancer_data)
tidy(cancer_anova)
```

`statistic` is $F$ and `p.value` is the p-value. $F = 6.70$ says the variation
between the group means is about 7 times the variation within the groups.

## Step 5. Confirm the p-value by hand

$F$ has $k - 1$ numerator and $N_{Total} - k$ denominator degrees of freedom, and
the p-value is always the **upper** tail:

```{r pf}
k <- 4
n_total <- nrow(cancer_data)
n_total

pf(6.699489, df1 = k - 1, df2 = n_total - k, lower.tail = FALSE)
```

# Part 4. After a significant ANOVA

A significant $F$ says the means are not all equal. It does not say which pair
differs. Two ways to follow up.

## Bonferroni

Divide $\alpha$ by the number of comparisons and use that as the bar for every
individual test:

```{r alpha-star}
alpha_star <- 0.05 / choose(4, 2)
alpha_star
```

Pull each group out as a vector, then run the six pairwise t tests:

```{r pairwise-tests}
both <- cancer_data %>% filter(treatment == "Both") %>% pull(tumor_volume)
neither <- cancer_data %>% filter(treatment == "Neither") %>% pull(tumor_volume)
can <- cancer_data %>% filter(treatment == "Cannabinoids") %>% pull(tumor_volume)
irr <- cancer_data %>% filter(treatment == "Irradiation") %>% pull(tumor_volume)

n_b <- t.test(neither, both)
c_b <- t.test(can, both)
i_b <- t.test(irr, both)
n_c <- t.test(neither, can)
n_i <- t.test(neither, irr)
c_i <- t.test(can, irr)

p_values <- c(n_b$p.value, c_b$p.value, i_b$p.value,
              n_c$p.value, n_i$p.value, c_i$p.value)
names(p_values) <- c("n_b", "c_b", "i_b", "n_c", "n_i", "c_i")
round(p_values, 4)
```

Which of them clear the adjusted bar?

```{r bonferroni-decision}
p_values < alpha_star
```

Only Neither vs. Both. Note that Irradiation vs. Both had p = 0.0413, which would
have passed at 0.05 but does **not** pass at 0.0083. That is the price of
protecting the overall error rate.

## Tukey's HSD

Tukey holds the error rate at 5% for the **whole family** of comparisons, so
`conf.level` stays at 0.95:

```{r tukey}
diffs <- TukeyHSD(cancer_anova, conf.level = 0.95) %>% tidy()
diffs %>% select(contrast, estimate, conf.low, conf.high, adj.p.value)
```

Two things to notice.

- The p-values are **adjusted**. Cannabinoids vs. Both was 0.054 unadjusted and
is 0.378 here.
- The intervals are adjusted too, so they are *simultaneous* 95% intervals for
the family. Interval and adjusted p-value therefore agree. An interval that
excludes 0 goes with an adjusted p-value below 0.05.

```{r tukey-decision}
diffs %>%
  mutate(excludes_zero = (conf.low > 0) | (conf.high < 0),
         significant = adj.p.value < 0.05) %>%
  select(contrast, excludes_zero, significant)
```

Tukey flags Irradiation vs. Both **and** Neither vs. Both, one more pair than
Bonferroni did. Tukey is more powerful here because it pools all four groups into
a single estimate of within-group variance with 18 degrees of freedom, while each
separate t test only had a handful.

## Visualising the pairwise differences

```{r tukey-plot}
ggplot(diffs, aes(x = contrast, y = estimate)) +
  geom_point() +
  geom_segment(aes(y = conf.low, yend = conf.high, xend = contrast)) +
  geom_hline(aes(yintercept = 0), lty = 2) +
  geom_text(aes(label = paste0("p-value:\n ", round(adj.p.value, 3))),
            nudge_x = 0.35, size = 3.5) +
  labs(y = "Estimated difference", x = "") +
  scale_x_discrete(expand = expansion(add = c(0.5, 0.9))) +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

Any interval that misses the dashed line at 0 is a difference Tukey calls real.

# Practice problems

**1.** You compare 5 groups. How many pairwise comparisons is that, and what
Bonferroni-adjusted $\alpha$ would you use to keep the overall type I error rate
at 0.05?

**2.** An ANOVA on 5 groups with 8 observations each returns $F = 3.1$. How many
degrees of freedom does the numerator have? The denominator? What is the p-value?

**3.** Four groups have sample standard deviations 3.1, 4.0, 5.8, and 7.2. Does
condition 3 hold?

**4.** A Tukey comparison reports an estimated difference of 12.4 with a 95%
simultaneous interval of 1.8 to 23.0. Without looking at the p-value column, is
this comparison significant at the family-wise 5% level?

## Answers

**1.**

```{r ans-1}
n_comparisons <- choose(5, 2)
n_comparisons

0.05 / n_comparisons
```

**2.** $k = 5$ so the numerator has $k - 1 = 4$ df. $N_{Total} = 5 \times 8 = 40$
so the denominator has $N_{Total} - k = 35$ df.

```{r ans-2}
pf(3.1, df1 = 4, df2 = 35, lower.tail = FALSE)
```

p is about 0.028, below 0.05, so at least one mean differs.

**3.**

```{r ans-3}
sds <- c(3.1, 4.0, 5.8, 7.2)
max(sds) / min(sds)
```

The ratio is about 2.3, above the rule of thumb of 2, so condition 3 is
questionable. With equal sample sizes of a decent size ANOVA is fairly forgiving,
but flag it.

**4.** Yes. The Tukey interval is already family-wise adjusted, so an interval
that excludes 0 corresponds to an adjusted p-value below 0.05. This is **not**
true of an unadjusted interval, so only reason this way when the interval and the
p-value come from the same adjusted procedure.
