---
title: "L25: The chi-squared goodness of fit"
output: html_document
---

```{r load-libraries, include=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
knitr::opts_chunk$set(fig.align = "center", message = FALSE, warning = FALSE)
options(width = 90)
```

Every calculation from L25 in one runnable place. Up to now a categorical variable
had **two** categories and we tested a single proportion. Today the variable has
**many** categories, and we ask one question about the whole distribution at once.

There is no external data file. Every number below is typed in from the slides.
Try the practice problems at the bottom before checking the answers.

## The test statistic

$$\chi^2 = \sum_{i=1}^k \frac{(O_i - E_i)^2}{E_i}$$

Three things to keep straight.

| Symbol | What it is | Common mistake |
|--------|-----------|----------------|
| $O_i$ | observed **count** in category $i$ | using a proportion instead of a count |
| $E_i$ | expected **count** under $H_0$, i.e. $n \times p_i$ | forgetting to multiply by $n$ |
| $df$ | $k - 1$, where $k$ is the number of categories | using $n - 1$ |

Squaring the numerator is what stops positive and negative gaps from cancelling.
Dividing by $E_i$ is what makes a gap of 100 matter more in a small category than
in a large one.

## Example 1: jury selection

A county's jury pool of 2500 people, against the racial composition of the county.

```{r jury-data}
race     <- c("White", "Black", "Latinx", "Asian", "Other")
observed <- c(1920, 347, 19, 84, 130)
pop_prop <- c(.422, .103, .251, .171, .053)

expected <- 2500 * pop_prop
expected
```

Note that the expected counts sum to the same total as the observed counts. That is
always true, and it is a free way to catch an arithmetic slip.

```{r jury-totals}
sum(observed)
sum(expected)
```

### Build the statistic by hand

```{r jury-by-hand}
contributions <- (observed - expected)^2 / expected
round(contributions, 3)

chi_sq <- sum(contributions)
chi_sq
```

The contributions are the most useful output here. A chi-square test gives one
p-value for the whole table, but the contributions tell you **which** categories
drove it. White, Latinx and Asian dominate.

```{r jury-pvalue}
pchisq(q = chi_sq, df = 4, lower.tail = FALSE)
```

R rounds this to 0. We always use `lower.tail = FALSE` because only *large* values
of $\chi^2$ are evidence against the null, so the p-value is always an upper tail area.

### The one-line version

```{r jury-chisq}
# x = observed counts, p = probabilities under the null
chisq.test(x = observed, p = pop_prop)
```

Same statistic, same degrees of freedom.

### The picture

```{r jury-plot, out.width="80%", fig.width=8, fig.height=3.6}
jury_data <- data.frame(race, observed, expected)

ggplot(jury_data, aes(y = race, x = observed)) +
  geom_segment(aes(xend = expected, yend = race)) +
  geom_point(aes(col = "Observed"), size = 2.5) +
  geom_point(aes(col = "Expected", x = expected), size = 2.5) +
  theme_minimal(base_size = 12) +
  labs(x = "Number of jurors", y = "Race/ethnicity", col = NULL,
       title = "Observed vs. expected number of jurors")
```

The three longest lines are exactly the three biggest contributions above. A good
picture usually tells you the answer before the test does.

## The chi-square distribution

```{r chisq-shapes, out.width="80%", fig.width=7, fig.height=3.6}
df_levels <- c("df=2", "df=4", "df=9", "df=15")

ggplot(data = data.frame(x = c(0.1, 30)), aes(x)) +
  stat_function(fun = dchisq, n = 201, args = list(df = 2),
                aes(col = factor("df=2", df_levels))) +
  stat_function(fun = dchisq, n = 201, args = list(df = 4),
                aes(col = factor("df=4", df_levels))) +
  stat_function(fun = dchisq, n = 201, args = list(df = 9),
                aes(col = factor("df=9", df_levels))) +
  stat_function(fun = dchisq, n = 201, args = list(df = 15),
                aes(col = factor("df=15", df_levels))) +
  coord_cartesian(ylim = c(0, 0.25)) +
  theme_minimal(base_size = 12) +
  labs(x = NULL, y = "Density", col = NULL, title = "Chi-squared distributions")
```

The distribution is never negative, and as $df$ rises the whole curve slides right.
That means with more categories you expect a larger statistic from ordinary noise
alone, so it takes a larger value to surprise you.

## Example 2: Alameda County vaccine doses

```{r alameda}
doses <- c(209299, 5024, 2012, 6617, 77668, 194647, 45509)
pop_p <- c(.306, .009, .011, .054, .187, .323, .110)

chisq.test(x = doses, p = pop_p)
```

The statistic is enormous, but so is the sample. With over half a million people,
even tiny departures from the null become statistically significant. So the honest
conclusion is that the disparity is **real**, not that it is large. To say how large,
look at the gaps themselves.

```{r alameda-gaps}
data.frame(
  group      = c("White", "PI", "Native Am.", "Multirace", "Latinx", "Asian", "AA/Black"),
  pct_pop    = pop_p * 100,
  pct_doses  = round(doses / sum(doses) * 100, 1)
)
```

## Example 3: cheating at dice?

Three dice, 100 rolls, counting how many sixes come up each time. Here the null is
not handed to us. We derive it from the binomial.

```{r dice-expected}
Expected <- dbinom(0:3, size = 3, prob = 1/6)
Expected

# as counts, out of 100 rolls
round(100 * Expected, 2)
```

```{r dice-chisq}
observed_dice <- c(47, 35, 15, 3)
chisq.test(x = observed_dice, p = Expected)
```

**Read the warning.** It is not noise. Look back at the expected counts: the "3 sixes"
cell expects only 0.46, which breaks the rule that every cell needs $E_i > 1$.

### Fixing it by collapsing categories

The standard fix is to combine sparse categories. Here we merge "2 sixes" and
"3 sixes" into a single "2 or more" category.

```{r dice-collapsed}
p_2plus   <- 1 - dbinom(0, 3, 1/6) - dbinom(1, 3, 1/6)
Expected2 <- c(dbinom(0:1, size = 3, prob = 1/6), p_2plus)
round(100 * Expected2, 2)   # all now above 5

chisq.test(x = c(47, 35, 18), p = Expected2)
```

No warning this time. The statistic is smaller and the degrees of freedom dropped to
2, which is what you should expect after throwing away detail, but the conclusion
survives. Always check the conditions, fix the problem, then see whether your answer
was ever really in danger.

## Conditions, in one place

- Fixed $n$, independent observations, mutually exclusive categories.
- $E_i \geq 5$ for at least 80% of cells.
- $E_i > 1$ for **every** cell.

All of these are about **expected** counts, never observed counts.

---

# Practice problems

### Problem 1

A die is rolled 120 times and lands on 1 through 6 this many times:
25, 17, 15, 23, 24, 16. Test whether the die is fair.

*Hint: under a fair die every face has probability 1/6, so all six expected counts
are the same.*

```{r p1}
rolls <- c(25, 17, 15, 23, 24, 16)
chisq.test(x = rolls, p = rep(1/6, 6))
```

**Answer.** $\chi^2 = 5.0$ on 5 degrees of freedom, p-value about 0.42. No evidence
against fairness. Note the expected count is $120/6 = 20$ in every cell, comfortably
above 5.

### Problem 2

For the die in Problem 1, compute the contribution of each face to the statistic by
hand and confirm they sum to what `chisq.test` reported.

```{r p2}
expected_fair <- rep(120/6, 6)
contrib <- (rolls - expected_fair)^2 / expected_fair
round(contrib, 4)
sum(contrib)
```

**Answer.** They sum to 5.0, matching the test exactly. Faces 1 and 3 tie for the
largest contribution at 1.25 each, being the two furthest from 20, but neither is
anywhere near large enough to matter.

### Problem 3

A clinic expects its patients to arrive 20% Monday, 20% Tuesday, 20% Wednesday,
20% Thursday and 20% Friday. One week it sees 30, 22, 18, 25 and 5 patients.
Is the distribution as expected? Check the conditions first.

```{r p3}
patients <- c(30, 22, 18, 25, 5)
n <- sum(patients)
expected_days <- n * rep(0.2, 5)
expected_days          # check all >= 5 before testing

chisq.test(x = patients, p = rep(0.2, 5))
```

**Answer.** The expected counts are all 20, so the conditions are fine even though
one *observed* count is only 5. Remember the rule is about expected counts. The test
gives $\chi^2 = 17.9$ on 4 degrees of freedom, p-value about 0.0013, so there is
strong evidence the arrivals are not evenly spread. Friday is the culprit, on its
own contributing $(5-20)^2/20 = 11.25$ of the 17.9.

### Problem 4

Why is the p-value for a chi-square test always an upper tail probability?

**Answer.** Because every departure from the null, in any direction and in any
category, makes the numerator $(O_i - E_i)^2$ larger and therefore makes $\chi^2$
larger. A small $\chi^2$ means the data look *like* the null, so only large values
are evidence against it. There is nothing a one-sided or two-sided distinction could
mean here.
