---
title: 'Lecture 13: The Poisson distribution'
---


```{r load-libraries, include=FALSE, message=F, warning=FALSE}
library(grid)
library(ggplot2)
```


## Recap: distributions so far

### Recap question 1: Normal

The heights of young women are approximately $N(\mu = 64.5, \sigma = 2.5)$ inches.
Using the 68-95-99.7 rule, about what proportion of young women are between
62 and 67 inches tall?

```{r recap-q1}
# 62 and 67 are exactly 1 SD below and above the mean, so about 68%
pnorm(67, mean = 64.5, sd = 2.5) - pnorm(62, mean = 64.5, sd = 2.5)
```

### Recap question 2: Binomial

A bottling line contaminates each bottle independently with probability
$p = 0.1$. In a box of $n = 10$ bottles, $X \sim \text{Bin}(10, 0.1)$.
What are the mean and standard deviation of $X$?

```{r recap-q2}
n <- 10; p <- 0.1
mu <- n * p                  # mean = np
sigma <- sqrt(n * p * (1-p)) # sd = sqrt(np(1-p))
c(mean = mu, sd = sigma)
```

## Poisson probabilities

### Probability distribution of a Poisson random variable

```{r, out.width="90%", echo=F, warning=F, message=FALSE}
#students you do NOT need to know this code
library(ggplot2)
point.probs <- rep(NA, 21)
point.probs2 <- rep(NA, 21)
point.probs3 <- rep(NA, 21)
point.probs4 <- rep(NA, 21)
point.probs5 <- rep(NA, 21)

for(i in 1:20) {
  point.probs5[i] <- dpois(x = i-1, lambda = 0.1)
  point.probs4[i] <- dpois(x = i-1, lambda = 0.3)
  point.probs[i] <- dpois(x = i-1, lambda = 1)
  point.probs2[i] <- dpois(x = i-1, lambda = 4)
  point.probs3[i] <- dpois(x = i-1, lambda = 10)
}

probs <- data.frame(point.probs, point.probs2, point.probs3, point.probs4)

plot1 <- ggplot(probs, aes(x = 0:20, y = point.probs)) +
  geom_col(fill = "pink", alpha = 0.5, col = "black") +
  labs(y = "Probability",
       x = "Number of events") +
  theme_minimal(base_size = 11) +
  scale_y_continuous(limits = c(0, 0.91)) +
  scale_x_continuous(breaks = c(0, 5, 10, 15, 20)) +
  ggtitle("lambda = 1")

plot2 <- ggplot(probs, aes(x = 0:20, y = point.probs2)) +
  geom_col(fill = "green", alpha = 0.5, col = "black") +
  labs(y = "Probability",
       x = "Number of events") +
  theme_minimal(base_size = 11) +
  scale_y_continuous(limits = c(0, 0.91)) +
  scale_x_continuous(breaks = c(0, 5, 10, 15, 20)) +
  ggtitle("lambda = 4")

plot3 <- ggplot(probs, aes(x = 0:20, y = point.probs3)) +
  geom_col(fill = "blue", alpha = 0.5, col = "black") +
  labs(y = "Probability",
       x = "Number of events") +
  theme_minimal(base_size = 11) +
  scale_y_continuous(limits = c(0, 0.91)) +
  scale_x_continuous(breaks = c(0, 5, 10, 15, 20)) +
  ggtitle("lambda = 10")

plot4 <- ggplot(probs, aes(x = 0:20, y = point.probs4)) +
  geom_col(fill = "yellow", alpha = 0.5, col = "black") +
  labs(y = "Probability",
       x = "Number of events") +
  theme_minimal(base_size = 11) +
  scale_y_continuous(limits = c(0, 0.91)) +
  scale_x_continuous(breaks = c(0, 5, 10, 15, 20)) +
  ggtitle("lambda = 0.3")

plot5 <- ggplot(probs, aes(x = 0:20, y = point.probs5)) +
  geom_col(fill = "forest green", alpha = 0.5, col = "black") +
  labs(y = "Probability",
       x = "Number of events") +
  theme_minimal(base_size = 11) +
  scale_y_continuous(limits = c(0, 0.91)) +
  scale_x_continuous(breaks = c(0, 5, 10, 15, 20)) +
  ggtitle("lambda = 0.1")


library(patchwork)

plot5 + plot4 + plot1 + plot2 + plot3 + plot_layout(ncol = 3)

```

## R functions for the Poisson

### Poisson distribution in R
In R the Poisson distribution is represented by `pois`

When working with the binomial distribution, we used `dbinom` and `pbinom`
to calculate probabilities.

With the Poisson distribution we will use `dpois` and `ppois`


In R, the function to calculate $P(X=x)$ for a Poisson is `dpois(x=?, lambda=?)`,
where `lambda` ($\lambda$) is equal to the average $\mu$ (which is the book's notation).


More detail on syntax is [*here*](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Poisson.html)

### Example: Mumps

In Iowa, the average number of reported cases of mumps per month is 0.1.
If we assume that cases of mumps are random and independent, the number $X$ of
monthly mumps cases in Iowa has approximately a Poisson distribution with
$\mu=0.1$. The probability that in a given month there is no more than 1 mumps
case is:

$P(X \leq 1) = P(X=0) + P(X=1)$


$$P(X=k)={\frac{e^{-\mu} \mu^k}{k!} }$$


### Example: Mumps


$P(X \leq 1) = P(X=0) + P(X=1)$


$$P(X=k)={\frac{e^{-\mu} \mu^k}{k!} }$$

= ${\frac{e^{-0.1} 0.1^0}{0!} } + {\frac{e^{-0.1} 0.1^1}{1!} }$ (note that 0! = 1, by definition, and $x^0$=1, for any value of x.)

= 0.9048 + 0.0905 = 0.9953

Thus, we expect to only see 0 cases 90.5% of the months, and 1 case 9.05% of the
time.

### Example: Mumps calculated using R using `ppois()` and `dpois()`

```{r}
dpois(x = 0, lambda = 0.1) + dpois(x = 1, lambda = 0.1)

# or,

ppois(q = 1, lambda = 0.1) # notice that lambda is the parameter
```

### Example: Mumps, continued

Suppose you saw 4 cases of Mumps in a given month. What are the chances of seeing
4 or more cases in any given month?

```{r}
# careful, we used q = 3 here. Why 3 and not 4?
1 - ppois(q = 3, lambda = 0.1)
```

Could you have performed this calculation using `dpois()`?

If you saw 4 or more cases in any given month, this is very unlikely under this
model. This suggests a substantial departure from the model, suggesting a
contagious outbreak (no longer independent)

### Example: Polydactyly

In the US, 1 in every 500 babies is born with an extra finger or toe. These
events are random and independent. Suppose that the local hospital delivers an
average of 268 babies per month. This means that for each month we expect to see
0.536 babies born with an extra finger or toe at that hospital (how do you calculate
0.536 here?). Let $X$ be the count of babies born with an extra finger or toe in a
month at that hospital.

a) What values can $X$ take?
b) What distribution might $X$ follow?
c) Give the mean and standard deviation of $X$.

### Example: Polydactyly, continued

To get a sense of what the data might look like, use R to simulate data across
five years (60 months) for this hospital.

```{r}
set.seed(200)
polydactyly <- rpois(n = 12*5, lambda = 0.536)
polydactyly
```

### Example: Polydactyly, continued
```{r, out.width="55%", fig.align='center'}
hist(polydactyly)
```

### Example: Polydactyly, continued
```{r}
mean(polydactyly)
sd(polydactyly)
```

### More random number generation

Examining a stream of Poisson-distributed random numbers helps us get a sense
of what these data look like. Can you think of a variable that might be Poisson-
distributed according to one of these distributions?

```{r}
rpois(100, lambda = 0.1)
rpois(100, lambda = 1)
```
