---
title: "L12:  The Binomial distribution student code"
---

<!-- libraries -->
```{r,include=FALSE,purl=FALSE}
library(knitr) # for include_graphics() 
library(dplyr)
library(tidyverse)

#don't need to worry about knowing these libraries
library(jpeg)
library(grid)
```

### Binomail Probability Distributions

We can use the formulas as shown to calculate the probability of a given number of successes from a binomial by hand

Or we could use R with 'dbinom(#successes,size,probability of success)'

This function calculates the probability of observing `x` successes when $X \sim Binom(n,p)$

```{r user, echo=TRUE}
dbinom(1,size=2,prob=0.29)
```


## Example trial of size 10

### Sampling distribution of a count in R

First, set up a large population of size 10,000 where 10% of the containers are 
contaminated by benzene. We call `benzene` a "success" since it is coded as 1.
We can see that 10% of the containers are contaminated and 1000 bottles are 
"successes"

We simulate these data:
```{r simulate-binomial-population}
container.id <- 1:10000
benzene <- c(rep(0, 9000), rep(1, 1000))
pop_data <- data.frame(container.id, benzene)
```

### Sampling distribution of a count in R

```{r}
# Calculate the population number of bottles contaminated by benzene and the 
# population mean proportion
pop_stats <- pop_data %>% summarize(pop_num_successes = sum(benzene),
                                    pop_mean = mean(benzene))

pop_stats
```

### Sampling distribution of a count in R

Take a sample of size 10 from the population. Note that 10 is much smaller than
10,000. 

- How many contaminated bottles are we expecting in the sample?
- Given that we sample 10, what is the full range of possible values we could 
see for X, the number of successes and $p$ the proportion of successes? 
- Which values are most likely?

```{r}
# first sample
set.seed(1)
sample_data <- pop_data %>% sample_n(10)
sample_data %>% summarize(sample_num_successes = sum(benzene),
                          sample_mean = mean(benzene))
```

### Sampling distribution of a count in R

We only took one sample, and got 2 successes for a sample mean of 20%. Is that 
usual or unusual? 

To see what is most likely, we need to imagine repeatedly taking samples of 
size 10 from the population and calculating the sample number of successes and 
proportion of successes for each sample.


For the next few slides, we focus on the sampling distribution for X.

### Sampling distribution of a count in R

The embedded code takes 1000 samples each of size 10. 

It then calculates the mean sample proportion and number of successes for each sample and stores all the results in a data frame. 

You don't need to know how the code works.

```{r take-many-samples, echo=FALSE}
# students, you don't need to understand the new parts of this code (i.e., how
# to write a function, or use `replicate`, `lapply`, or `bind_rows`)

calc_sample_stats <- function(df) {
  df %>% 
    summarize(sample_proportion = mean(benzene),
              sample_num_successes = sum(benzene))
}

many.sample.stats <- replicate(1000, sample_n(pop_data, 10), simplify = F) %>%
  lapply(., calc_sample_stats) %>%
  bind_rows() %>%
  mutate(sample.id = 1:n())
```

### Sampling distribution of a count in R

Here are the first rows of the data frame we made on the previous slide.
Each row represents an independent sample from the population. 

```{r show-many-samples}
head(many.sample.stats)
```

### Sampling distribution of a count in R

We want to know: Of the 1000 samples, what percent observed 0 contaminated 
bottles? What percent observed 1 contaminated bottle? And so on. We can used 
`dplyr` functions to calculate this and plot the results in a histogram.

```{r histogram-sampling-distribution, warning=F}
aggregated.stats <- many.sample.stats %>% 
  group_by(sample_num_successes) %>% 
  summarize(percent = n()/1000)
```

### Sampling distribution of a count in R
```{r hist-sample, warning=F}
aggregated.stats
```

### Sampling distribution of a count in R

```{r, fig.align='center', out.width="80%", warning=F, echo = F}
# note: We want a histogram, but R is interpreting these data as a bar chart
# this is, we don't want space between the bars because the x variable is quantitative. 

ggplot(aggregated.stats, aes(x = as.numeric(sample_num_successes), y = percent)) + 
  geom_histogram(col = "grey", stat = "identity") +
  theme_minimal(base_size = 15) + 
  labs(title = "Histogram of the number of \nsuccesses observed across 1000 samples",
       x = "Number of contaminated bottles", y = "percent in each bin") +
  scale_x_continuous(labels = 0:5, breaks = 0:5)
```


### Worked probabilities, x = 1

We can check our calculations using the `dbinom()` function in R. 

```{r}
dbinom(x = 1, size = 10, prob = 0.1)
```


### Worked probabilities, x = 2
Note: we can get our calculators or R to perform this calculation for us. On our calculator, 
we need the button ${n \choose k}$, pronounced "n choose k", and asks how many ways
are there to have $k$ successes when there are $n$ individuals? In R we need the 
function `choose(n, k)`

```{r 10-choose-2}
choose(10, 2)
```

There are 45 ways to observe exactly two contaminated bottles when you 
have ten bottles observed.

**Make sure you can also perform this calculation on your calculator!**



### All of the combinations with 10 bottles 
Each of these is written as ${10 \choose k}$, where k is 0, 1, 2, ..., 10. This is
known as the **binomial coefficient**.

Let's compute `choose(n, k)`, for n=10, and k=0, 1, 2, ..., 10:
```{r 10-choose-k, echo=F}
choices<- cbind(choose(10, 0),
choose(10, 1),
choose(10, 2),
choose(10, 3),
choose(10, 4),
choose(10, 5),
choose(10, 6),
choose(10, 7),
choose(10, 8),
choose(10, 9),
choose(10, 10))
choices
```

## Binomial probability in R

### Binomial probability in R using `dbinom()` 

- Recall for Normal distributions we used `pnorm()` to calculate the probability
**below** a given number.
- For discrete distributions we can calculate the probability of observing a 
specific value. For example, we can ask: What is the probability that exactly
3 of the ten bottles were contaminated when the risk of contamination was 10%?

- `dbinom()` is used to compute *exactly* 3

```{r}
dbinom(x = 3, size = 10, prob = 0.1)
```

### Binomial probability in R using  `pbinom()`

- Recall for Normal distributions we used `pnorm()` to calculate the probability
**below** a given number.

- For our Binomial, we can also ask, what is the probability that 3 or less of the ten bottles were
contaminated when the risk of contamination was 10%?
-  `pbinom()` is used to compute 3 *or less*

```{r}
dbinom(x = 3, size = 10, prob = 0.1)
pbinom(q = 3, size = 10, prob = 0.1)
```


### Histogram of binomial probabilities

This histogram shows the probability of observing each value of $X$. That is, it
shows the $P(X =x)$, for $x$ in 0,1,2, ... 10, when $X \sim Binom(n=10, p = 0.1)$

```{r, out.width="80%", fig.align='center', echo=F, warning=F}
#students, don't worry about this code
point.probs <- rep(NA, 11)

for(i in 1:11) {
  point.probs[i] <- dbinom(x = i-1, size = 10, prob = 0.1)
}

ggplot(data.frame(point.probs), aes(x = 0:10, y = point.probs)) +
  geom_histogram(stat = "identity", binwidth = 1) +
  labs(y = "Probability",
       x = "Number of successes") +
  theme_minimal(base_size = 15) +
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8, 10))

```

### Exact discrete probability, graphed

```{r, fig.align='center', out.width="80%", echo=F, warning=F}
ggplot(data.frame(point.probs), aes(x = 0:10, y = point.probs)) +
  geom_histogram(stat = "identity", binwidth = 1, aes(fill = as.factor(round(point.probs, 2) == 0.06 ))) +
  labs(y = "Probability",
       x = "Number of successes") +
  theme_minimal(base_size = 15) +
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8, 10)) + 
  scale_fill_discrete(labels = c("other", "P(X=3)"), breaks = c(FALSE, TRUE), name = "")
```

```{r exact-prob}
dbinom(x = 3, size = 10, prob = 0.1)
```


### Cumulative discrete probability, graphed

```{r, fig.align='center', out.width="80%", warning=F, echo=F}
ggplot(data.frame(point.probs), aes(x = 0:10, y = point.probs)) +
  geom_histogram(stat = "identity", binwidth = 1, aes(fill = as.factor(round(point.probs, 2) >= 0.06 ))) +
  labs(y = "Probability",
       x = "Number of successes") +
  theme_minimal(base_size = 15) +
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8, 10)) + 
  scale_fill_discrete(labels = c("other", "P(X<=3)"), breaks = c(FALSE, TRUE), name = "")
```

```{r}
pbinom(q = 3, size = 10, prob = 0.1)
```



## Normal approximation of a binomial

### Histogram of binomial probabilities with different values for p

Here we have n = 10, and p = 0.10 (green), 0.5 (pink), and 0.8 (blue)

```{r, out.width="80%", fig.align='center', echo=F, warning=F}
#students, don't worry about this code
point.probs <- rep(NA, 11)
point.probs2 <- rep(NA, 11)
point.probs3 <- rep(NA, 11)

for(i in 1:11) {
  point.probs[i] <- dbinom(x = i-1, size = 10, prob = 0.5)
  point.probs2[i] <- dbinom(x = i-1, size = 10, prob = 0.10)
  point.probs3[i] <- dbinom(x = i-1, size = 10, prob = 0.80)
}

probs <- data.frame(point.probs, point.probs2, point.probs3)

plot1 <- ggplot(probs, aes(x = 0:10, y = point.probs)) +
  geom_histogram(stat = "identity", binwidth = 1, fill = "pink", alpha = 0.5, col = "black") +
  labs(y = "Probability",
       x = "Number of successes") +
  theme_minimal(base_size = 15) +
  scale_y_continuous(limits = c(0, 0.4)) +
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8, 10))

plot2 <- ggplot(probs, aes(x = 0:10, y = point.probs2)) +
  geom_histogram(stat = "identity", binwidth = 1, fill = "green", alpha = 0.5, col = "black") +
  labs(y = "Probability",
       x = "Number of successes") +
  theme_minimal(base_size = 15) +
  scale_y_continuous(limits = c(0, 0.4)) +
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8, 10))

plot3 <- ggplot(probs, aes(x = 0:10, y = point.probs3)) +
  geom_histogram(stat = "identity", binwidth = 1, fill = "blue", alpha = 0.5, col = "black") +
  labs(y = "Probability",
       x = "Number of successes") +
  theme_minimal(base_size = 15) +
  scale_y_continuous(limits = c(0, 0.4)) + 
  scale_x_continuous(breaks = c(0, 2, 4, 6, 8, 10))


library(patchwork)

plot2 + plot1 + plot3 + plot_layout(ncol = 3)
```


### An approximation to the binomial distribution when n is large

Consider the probability distribution for $P(X=k)={2000\choose k}0.62^k(1-0.62)^{2000-k}$

What shape does this remind you of?

```{r calculate-prob-distn-binom, echo = F, fig.align='center', out.width="80%", warning=F}
#students, don't worry about this code
point.probs.2k <- rep(NA, 2001)

for(i in 1:2001) {
  point.probs.2k[i] <- dbinom(x = i-1, size = 2000, prob = 0.62)
}

zoomed.out <- ggplot(data.frame(point.probs.2k), aes(x = 0:2000, y = point.probs.2k)) +
  geom_histogram(stat = "identity", binwidth = 1) +
  labs(y = "Probability",
       x = "Number of successes") +
  theme_minimal(base_size = 15) 

zoomed.in <- ggplot(data.frame(point.probs.2k), aes(x = 0:2000, y = point.probs.2k)) +
  geom_histogram(stat = "identity", binwidth = 1, col = "white") +
  labs(y = "Probability",
       x = "Number of successes") +
  theme_minimal(base_size = 15) +
  scale_x_continuous(limits = c(1150, 1350))

#zoomed.out + zoomed.in + plot_layout(ncol = 2)

zoomed.in
```


### How much data is within 1 SD of the mean?

1240 +/- 1 SD gives the range {1218.293, 1261.707}

Thus, we can use R to add up all the probabilities between X = 1218 and X = 1262
to give an approximate guess to the area 1 SD from the mean:

This code cycles through the probabilities to add them up

```{r}
#students, no need to know how to write this code.
cumulative.prob <- 0
 
for(i in 1218:1262){
  cumulative.prob <- cumulative.prob + point.probs.2k[i]
}

cumulative.prob
```

### How much data is within 2 SD of the mean?

1240 +/- 2 SD gives the range {1196.586, 1283.414}

Thus, we can use R to add up all the probabilities between X = 1197 and X = 1283
to give an approximate guess to the area 1 SD from the mean:

This code cycles through the probabilities to add them up

```{r}
#students, no need to know how to write this code.
cumulative.prob.2 <- 0
 
for(i in 1197:1283){
  cumulative.prob.2 <- cumulative.prob.2 + point.probs.2k[i]
}

cumulative.prob.2
```
- You could also perform the check for 3 SD


### Example calculation of the Normal approximation to the Binomial 

Suppose we want to calculate $P(X \geq 1250)$ using the Normal approximation.

```{r}
# write the Normal code
1- pnorm(q = 1250, mean = 1240 , sd = 21.70714)
```

Check how well the approximation worked: 

```{r}
# write the binomial code and see how well the approximation is
1 - pbinom(q = 1249, size = 2000, prob = 0.62)
```


