---
title: "Bonus material: Bootstraps and permutation tests"
output: html_document
---

```{r load-libraries, include=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(infer)
knitr::opts_chunk$set(fig.align = "center", message = FALSE, warning = FALSE)
```

Every calculation from the bonus lecture in one runnable place. Everything we have
done so far got a confidence interval or a p-value out of a **formula**. Both
methods here get the same quantities out of **resampling the data on a computer**
instead, so they keep working when the formula's assumptions do not.

The only external file you need is `hsb2.csv`, which should sit in the same folder
as this file. Everything else is typed in below. Try the practice problems at the
bottom before checking the answers.

If `library(infer)` fails, install it once with `install.packages("infer")`.

## The two methods side by side

| | Bootstrap | Permutation test |
|---|---|---|
| Answers | what is the **CI** for a parameter? | what is the **p-value**? |
| Resample | **with** replacement, from the sample | **without** replacement, reshuffling the group labels |
| Keeps intact | nothing, each draw is independent | the group sizes, and the pool of outcome values |
| Assumes | the sample is a SRS | the null of no association is true |
| Read off | the 2.5th and 97.5th percentiles | how often the reshuffles beat the observed statistic |

The one-sentence version. The bootstrap asks "how much would this statistic
bounce around if I could sample again?", and the permutation test asks "how
often would chance alone produce a gap this big?".

## Part 1: the bootstrap

### The serum data

Efron's serum measurements on 54 patients.

```{r}
serum_data <- data.frame(
  serum = c(0.1, 0.1, 0.2, 0.4, 0.4, 0.6, 0.8, 0.8, 0.9, 0.9,
            1.3, 1.3, 1.4, 1.5, 1.6, 1.6, 1.7, 1.7, 1.7, 1.8,
            2.0, 2.0, 2.2, 2.2, 2.2, 2.3, 2.3, 2.4, 2.4, 2.4,
            2.4, 2.4, 2.4, 2.5, 2.5, 2.5, 2.7, 2.7, 2.8, 2.9,
            2.9, 2.9, 3.0, 3.1, 3.1, 3.2, 3.2, 3.3, 3.3, 3.5,
            4.4, 4.5, 6.4, 9.4))

ggplot(serum_data, aes(serum)) +
  geom_histogram(binwidth = 0.5, col = "white") +
  theme_minimal(base_size = 13) +
  labs(title = "Histogram of serum measurements")
```

That long right tail is the whole reason this lecture exists. A t-interval leans
on the CLT, and here the population is clearly not Normal.

### The theory-based CI for the mean, two ways

```{r}
t_star <- qt(0.975, df = 53)  # t* for a 95% CI

serum_data %>%
  summarise(mean_serum = mean(serum),
            se_serum = sd(serum) / sqrt(n()),
            lower_CI = mean_serum - t_star * se_serum,
            upper_CI = mean_serum + t_star * se_serum)
```

```{r}
t.test(serum_data$serum)
```

Both give a 95% CI for $\mu$ of 1.90 to 2.74. Note this is an interval for the
**mean**, not the proportion and not the median.

### Bootstrap CI for the median

There is no formula for the standard error of a median, so we build the sampling
distribution by brute force. Resample 54 values **with replacement** from the 54
we have, take the median, and repeat 10,000 times.

```{r}
number_of_bootstraps <- 10000
sample_size <- 54

calc_sample_stats <- function(df) {
  df %>% summarize(sample_median = median(serum))
}

set.seed(1)   # so you get the same answer we did

many_sample_medians <- replicate(number_of_bootstraps,
                                 sample_n(serum_data, sample_size, replace = TRUE),
                                 simplify = FALSE) %>%
  lapply(., calc_sample_stats) %>%
  bind_rows() %>%
  mutate(sample.id = 1:n())

head(many_sample_medians)
```

You are not expected to be able to write that loop. You **are** expected to
understand the next chunk, which reads the interval straight off the percentiles.

```{r}
bounds <- many_sample_medians %>%
  summarise(lower_CI = quantile(sample_median, 0.025),
            upper_CI = quantile(sample_median, 0.975))

bounds
```

```{r}
ggplot(many_sample_medians, aes(x = sample_median)) +
  geom_histogram(binwidth = 0.07, col = "white", fill = "#99d8c9") +
  labs(x = "Bootstrapped medians") +
  geom_vline(xintercept = median(serum_data$serum)) +
  geom_vline(xintercept = bounds$lower_CI, lty = 2) +
  geom_vline(xintercept = bounds$upper_CI, lty = 2) +
  theme_minimal(base_size = 13)
```

Our estimate of the median is 2.35 and the bootstrap 95% CI is 1.9 to 2.5. Two
things to notice. The distribution is skewed left, and the interval is **not**
symmetric around 2.35. A formula of the form estimate $\pm$ margin of error could
never have produced that.

### Bootstrap CI for a correlation

Same recipe, different statistic. This is the point of the bootstrap: the recipe
does not change when the statistic does.

```{r}
hsb2 <- read.table("hsb2.csv", sep = ",", header = TRUE)

corr_math_write <- hsb2 %>%
  summarise(correlation = cor(write, math))
corr_math_write
```

```{r}
ggplot(hsb2, aes(write, math)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  annotate("text", x = 40, y = 70, label = "corr = 0.62") +
  theme_minimal(base_size = 13)
```

```{r}
number_of_bootstraps <- 1000
sample_size <- 200

calc_sample_stats <- function(df) {
  df %>% summarize(sample_corr = cor(math, write))
}

set.seed(1)

many_sample_correlations <- replicate(number_of_bootstraps,
                                      sample_n(hsb2, sample_size, replace = TRUE),
                                      simplify = FALSE) %>%
  lapply(., calc_sample_stats) %>%
  bind_rows() %>%
  mutate(sample.id = 1:n())

bounds <- many_sample_correlations %>%
  summarise(lower_CI = quantile(sample_corr, 0.025),
            upper_CI = quantile(sample_corr, 0.975))

bounds
```

The correlation is 0.62 with a 95% CI of 0.53 to 0.69. The interval excludes 0,
so writing and math scores are clearly related.

## Part 2: permutation tests

### The mosquito data

Volunteers were randomised to drink beer or water, and the number of mosquitoes
flying toward each of them was counted.

```{r}
beer <- c(27, 19, 20, 20, 23, 17, 21, 24, 31, 26, 28, 20,
          27, 19, 25, 31, 24, 28, 24, 29, 21, 21, 18, 27,
          20)

water <- c(21, 19, 13, 22, 15, 22, 15, 22, 20, 12, 24, 24,
           21, 19, 18, 16, 23, 20)

mosq_data <- data.frame(num_mosquitos = c(beer, water),
                        treatment = c(rep("beer", 25),
                                      rep("water", 18)))

ggplot(mosq_data, aes(x = treatment, y = num_mosquitos)) +
  geom_boxplot(aes(fill = treatment), show.legend = FALSE) +
  theme_minimal(base_size = 13) +
  labs(y = "Number of mosquitos", x = "")
```

### The parametric answer first

```{r}
t.test(beer, water, alternative = "two.sided")
```

### The same question by reshuffling

The null says the label "beer" tells you nothing about the count. If that is
true, we can rip the 43 labels off, shuffle them, staple them back on, and the
difference in means we get is just as legitimate as the one we observed. Do that
1000 times and you have the null distribution.

```{r}
set.seed(1)

null_distn <- mosq_data %>%
  specify(response = num_mosquitos, explanatory = treatment) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("beer", "water"))

nrow(null_distn)
head(as.data.frame(null_distn), 4)
```

The observed difference is the thing we compare against that distribution.

```{r}
obs_diff <- mean(beer) - mean(water)
obs_diff
```

```{r}
null_distn %>%
  visualize() +
  shade_p_value(obs_stat = obs_diff, direction = "two_sided")
```

```{r}
null_distn %>%
  get_p_value(obs_stat = obs_diff, direction = "two_sided")
```

Only about 2 reshuffles in 1000 landed as far from zero as 4.38, so p is about
0.002 and we reject the null. The t-test gave 0.0007. The two methods agree on
the conclusion, which is what you should expect when the sample is large enough
for the t-test to be trustworthy.

Note that a permutation p-value can never be exactly 0. If you see `0`, it means
"smaller than 1/reps", so run more reps before reporting it.

## Practice problems

### Problem 1

Change the bootstrap for the serum data so it gives a 95% CI for the **first
quartile** $Q_1$ instead of the median.

*Hint: only the function inside `calc_sample_stats()` needs to change.*

```{r}
set.seed(1)

calc_q1 <- function(df) {
  df %>% summarize(sample_q1 = quantile(serum, 0.25))
}

many_q1 <- replicate(10000,
                     sample_n(serum_data, 54, replace = TRUE),
                     simplify = FALSE) %>%
  lapply(., calc_q1) %>%
  bind_rows()

many_q1 %>%
  summarise(q1 = quantile(serum_data$serum, 0.25),
            lower_CI = quantile(sample_q1, 0.025),
            upper_CI = quantile(sample_q1, 0.975))
```

The recipe did not change at all. That is the answer to "which statistics can I
bootstrap?" — essentially any of them.

### Problem 2

You bootstrap a median 1000 times and get a 95% CI of 1.9 to 2.5. A classmate
says "so there is a 95% probability the true median is between 1.9 and 2.5."
What is wrong with that sentence?

*Hint: think back to how we interpreted the t-interval.*

The true median is a fixed number, not a random one, so it does not have a
probability of being anywhere. The 95% describes the **method**: if we repeated
the whole study and this procedure many times, 95% of the intervals we built
would cover the true median. Bootstrapping changes how we compute the interval,
not how we interpret it.

### Problem 3

Turn the 95% bootstrap CI for the correlation into a **99%** CI. Does it get
wider or narrower, and why?

```{r}
many_sample_correlations %>%
  summarise(lower_CI = quantile(sample_corr, 0.005),
            upper_CI = quantile(sample_corr, 0.995))
```

Wider. To be more confident of catching the true value you have to cast a bigger
net, so we take the middle 99% of the bootstrap distribution instead of the
middle 95%.

### Problem 4

In the mosquito permutation test, what would happen to the null distribution if
we shuffled the labels but kept 25 in the beer group and 18 in the water group,
versus splitting them 30 and 13? Why does `generate(type = "permute")` always
preserve the original group sizes?

Because the group sizes are part of the study design, not part of the null
hypothesis. The null claims the labels are unrelated to the counts, not that the
experiment enrolled different numbers of people. Holding the sizes fixed makes
the reshuffled differences comparable to the observed one, since a difference of
means based on 25 versus 18 has a different amount of noise than one based on 30
versus 13.

### Problem 5

Run the mosquito permutation test as a **one-sided** test of whether beer
drinkers attract *more* mosquitoes. Should the p-value be bigger or smaller than
the two-sided one?

```{r}
null_distn %>%
  get_p_value(obs_stat = obs_diff, direction = "greater")
```

Smaller, about half the two-sided value, because we now count only the reshuffles
in the upper tail rather than both tails. Choose the direction from the research
question **before** looking at the data, not after.
