---
title: "L22: Inference for linear regression"
output: html_document
---

```{r load-libraries, include=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(broom)
library(patchwork)
knitr::opts_chunk$set(fig.align = "center", message = FALSE, warning = FALSE)
options(width = 90)
```

Every calculation from L22 in one runnable place. In Part I you fitted a line and
read off a slope. This lecture attaches uncertainty to that slope, using the same
machinery (standard error, t distribution, confidence interval, p-value) that you
have used all through Part III.

The only external file you need is `Ch03_Manatee-deaths.csv`, which should sit in
the same folder as this file. Everything else is created in the code below. Try
the practice problems at the bottom before checking the answers.

## The three broom functions

Keep these straight by what each one gives *back*:

| Function | Returns | Go there for |
|----------|---------|--------------|
| `tidy(your_lm)` | one row per coefficient | estimate, std.error, statistic, p.value |
| `glance(your_lm)` | one row for the model | `r.squared`, `sigma` |
| `augment(your_lm)` | one row per observation | `.fitted`, `.resid` (every diagnostic plot) |

# Part I recap: the manatees

Each row is one year. `powerboats` is the number of powerboats registered in
Florida that year (in thousands), and `deaths` is the number of manatees killed
by boats.

```{r manatee-data}
mana_data <- read_csv("Ch03_Manatee-deaths.csv")
head(mana_data)
```

```{r manatee-scatter, out.width="60%"}
mana_scatter <- ggplot(mana_data,
                       aes(x = powerboats, y = deaths)) +
  geom_point() +
  theme_minimal(base_size = 15)

mana_scatter
```

Correlation, then the line of best fit:

```{r manatee-model}
mana_cor <- mana_data %>%
  summarize(corr_mana = cor(powerboats, deaths))
mana_cor

mana_lm <- lm(deaths ~ powerboats, mana_data)
tidy(mana_lm)
glance(mana_lm)
```

- **Intercept** (-46.8): the predicted deaths when zero powerboats are
  registered. This is far outside the range of the data (which runs from 447 to
  1027 thousand boats), so do not read anything biological into it.
- **Slope** (0.136): one extra thousand boats registered is *associated with*
  about 0.136 extra manatee deaths that year.

With a single explanatory variable, r-squared really is the correlation squared:

```{r corr-vs-rsq}
mana_cor$corr_mana        # r
mana_cor$corr_mana^2      # r squared by hand
glance(mana_lm)$r.squared # r squared from the model
```

# The four assumptions

Inference on a regression is only valid if:

1. The relationship between x and y is **linear in the population**.
2. y varies **Normally** about the line, i.e. the residuals are Normal.
3. Observations are **independent** (this comes from the study design, no plot
   can check it).
4. The **standard deviation of the responses is the same for all values of x**.

Assumptions 1, 2 and 4 are checked with the residuals, where

$$\hat{e} = \text{observed} - \text{fitted} = y - (\hat{a} + \hat{b}x)$$

The four diagnostic panels, and what each is for:

| Panel | Plot | Checks |
|-------|------|--------|
| (a) | scatter plot with residual segments | linearity (assumption 1) |
| (b) | QQ plot of the residuals | Normality (assumption 2) |
| (c) | fitted values vs residuals | linearity **and** constant variance (1 and 4) |
| (d) | boxplot of y vs boxplot of residuals | not an assumption, this measures usefulness |

Panel (c) is the most useful of the four. A **smile or frown** means the truth was
curved. A **megaphone** means the spread is not constant.

## A reusable diagnostic function

The four panels are always built the same way, so write it once:

```{r diagnostic-function}
diag_plots <- function(model, caption = "") {
  aug <- augment(model)
  yvar <- names(aug)[1]   # outcome
  xvar <- names(aug)[2]   # predictor

  p1 <- ggplot(aug, aes(x = .data[[xvar]], y = .data[[yvar]])) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_point() +
    geom_segment(aes(xend = .data[[xvar]], yend = .fitted), lty = 2) +
    theme_minimal(base_size = 11) +
    labs(title = "(a) Scatter plot")

  p2 <- ggplot(aug, aes(sample = .resid)) +
    geom_qq() +
    geom_qq_line() +
    theme_minimal(base_size = 11) +
    labs(y = "Residuals", x = "Theoretical quantiles", title = "(b) QQplot")

  p3 <- ggplot(aug, aes(y = .resid, x = .fitted)) +
    geom_point() +
    geom_hline(yintercept = 0) +
    theme_minimal(base_size = 11) +
    labs(y = "Residuals", x = "Fitted values",
         title = "(c) Fitted vs. residuals")

  p4 <- aug %>%
    select(.resid, all_of(yvar)) %>%
    gather(key = "type", value = "value") %>%
    ggplot(aes(y = value)) +
    geom_boxplot(aes(fill = type)) +
    theme_minimal(base_size = 11) +
    labs(title = "(d) Amount explained")

  p1 + p2 + p3 + p4 + plot_layout(nrow = 2) +
    plot_annotation(caption = caption)
}
```

## Example 1: everything is fine

```{r diag-example-1, out.width="85%"}
set.seed(123)
x <- seq(1, 10, length.out = 60)
y <- 12 + 2.5 * x + rnorm(n = 60, mean = 0, sd = 2)
dat1 <- data.frame(x, y)

lm1 <- lm(y ~ x, data = dat1)
diag_plots(lm1, "A good fit to the data")
```

Residuals fall on both sides of the line with no trend, the QQ plot is straight,
panel (c) is a shapeless cloud, and the residual box is much shorter than the y
box. Inference is allowed, and the model is useful.

## Example 2: linearity is violated

```{r diag-example-2, out.width="85%"}
set.seed(456)
x2 <- seq(1, 10, length.out = 60)
# the truth is curved, but we fit a straight line to it
y2 <- 12 + 2.5 * x2 + 1.2 * x2^2 + rnorm(n = 60, mean = 0, sd = 3)
dat2 <- data.frame(x2, y2)

lm2 <- lm(y2 ~ x2, data = dat2)
diag_plots(lm2, "The linearity assumption does not hold")
```

Panel (a) looks passable, which is exactly why you must look at panel (c). That
**U shape** is the fingerprint of a straight line fitted to something curved. The
fix is a squared term, $\hat{y} = \hat{a} + \hat{b}x + \hat{c}x^2$.

Note panel (d): the wrong model still explains a lot of the variation. *Useful*
and *correctly specified* are two different questions.

## Example 3: constant variance is violated

```{r diag-example-3, out.width="85%"}
set.seed(789)
x3 <- seq(1, 10, length.out = 60)
# the sd of the noise grows with x
y3 <- 12 + 2.5 * x3 + rnorm(n = 60, mean = 0, sd = 2 * x3)
dat3 <- data.frame(x3, y3)

lm3 <- lm(y3 ~ x3, data = dat3)
diag_plots(lm3, "The constant variance assumption does not hold")
```

Panel (c) **fans out**. This matters because all of the inference below uses a
single number to describe the noise at every point on the line. When the noise is
small at one end and large at the other, that single number is wrong everywhere.

# The frog example

Twenty frogs. `temp` is air temperature in degrees Celsius, `freq` is mating
calls per minute.

```{r frog-data}
frog_data <- tibble(
  id   = 1:20,
  temp = c(19, 21, 22, 22, 23, 23, 23, 23, 23, 24,
           24, 24, 24, 25, 25, 25, 25, 26, 26, 27),
  freq = c(38, 42, 45, 45, 41, 45, 48, 50, 53, 51,
           48, 53, 47, 53, 49, 56, 53, 55, 55, 54))
```

```{r frog-scatter, out.width="60%"}
ggplot(frog_data, aes(x = temp, y = freq)) +
  geom_point() +
  theme_minimal(base_size = 15) +
  labs(x = "Temperature (Celsius)",
       y = "Frequency of mating calls")
```

```{r frog-model}
frog_lm <- lm(formula = freq ~ temp, data = frog_data)
tidy(frog_lm)
```

One extra degree Celsius is associated with about 2.33 extra calls per minute.

## Diagnostics for the frog model

`augment()` glues `.fitted` and `.resid` onto the original data frame:

```{r frog-augment}
frog_lm_augment <- augment(frog_lm)
head(frog_lm_augment)
```

Panel (a), the fitted line with the residuals drawn in:

```{r frog-panel-a, out.width="60%"}
frog_resid <- ggplot(frog_lm_augment, aes(temp, freq)) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(alpha = 0.5) +
  geom_segment(aes(xend = temp, yend = .fitted),
               lty = 2, alpha = 0.5) +
  theme_minimal(base_size = 15) +
  labs(title = "(a) Scatter plot")

frog_resid
```

Panel (b), the QQ plot. Note the aesthetic is `sample = .resid`:

```{r frog-panel-b, out.width="60%"}
frog_qq <- ggplot(frog_lm_augment, aes(sample = .resid)) +
  geom_qq() +
  geom_qq_line() +
  theme_minimal(base_size = 15) +
  labs(y = "Residuals", x = "Theoretical quantiles",
       title = "(b) QQplot")

frog_qq
```

Panel (c), fitted vs residuals:

```{r frog-panel-c, out.width="60%"}
frog_fitresid <- ggplot(frog_lm_augment,
                        aes(y = .resid, x = .fitted)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  theme_minimal(base_size = 15) +
  labs(y = "Residuals", x = "Fitted values",
       title = "(c) Fitted vs. residuals")

frog_fitresid
```

Panel (d), amount explained:

```{r frog-panel-d, out.width="60%"}
frog_gather <- frog_lm_augment %>%
  select(freq, .resid) %>%
  gather(key = "type", value = "value", freq, .resid)

frog_expl <- ggplot(frog_gather, aes(y = value)) +
  geom_boxplot(aes(fill = type)) +
  theme_minimal(base_size = 15) +
  labs(title = "(d) Amount explained")

frog_expl
```

No smile, no megaphone, the QQ plot is acceptable at n = 20, and the twenty frogs
are independent by design. Inference is justified.

# The regression standard error

$$s = \sqrt{\frac{1}{n-2}\sum_{i=1}^{n}{(y-\hat{y})^2}}$$

This is the standard deviation of the residuals. The denominator is $n-2$ because
fitting a line costs two degrees of freedom, one for the intercept and one for
the slope. R calls it `sigma`:

```{r frog-sigma}
glance(frog_lm)
```

`sigma` = 2.82 calls per minute. On average an actual frog's calling rate sits
about 2.8 calls away from what the line predicted for it. You can see the same
spread directly in the residuals:

```{r frog-resid-summary}
frog_lm_augment %>% summarise(min_resid  = min(.resid),
                              q25_resid  = quantile(.resid, 0.25),
                              mean_resid = mean(.resid),
                              q75_resid  = quantile(.resid, 0.75),
                              max_resid  = max(.resid))
```

The mean residual prints as something times $10^{-15}$, which is zero. Residuals
from a fitted line always sum to zero by construction.

# Hypothesis test for the slope

$$H_0: b = 0 \qquad H_a: b \neq 0$$

A slope of zero is a **flat line**: knowing x tells you nothing about y. The test
statistic is signal over noise,

$$t = \frac{\hat{b}}{SE_b}, \qquad SE_b = \frac{s}{\sqrt{\sum(x-\bar{x})^2}}$$

with $n-2$ degrees of freedom. All four pieces are already in `tidy()`:

```{r frog-test}
tidy(frog_lm)
```

- `estimate` is $\hat{b}$ = 2.33 (the signal)
- `std.error` is $SE_b$ = 0.347 (the noise)
- `statistic` is $t = 2.330816 / 0.3467893 = 6.72$
- `p.value` is the two sided p-value, on $n - 2 = 18$ degrees of freedom

The p-value is an ordinary t distribution tail, doubled because the alternative
is two sided:

```{r frog-pvalue}
pt(q = 6.7211302, df = 18, lower.tail = FALSE) * 2
```

We reject $H_0$. There is strong evidence of an association between temperature
and calling frequency.

# Confidence interval for the slope

$$\hat{b} \pm t^{*}SE_b$$

`qt()` works with area *to the left*, so for a 95% interval you enter 0.975, not
0.95 (2.5% is left in each tail):

```{r frog-ci}
t_star <- qt(p = 0.975, df = 18)
t_star

2.330816 + c(-1, 1) * t_star * 0.3467893
```

Or let R do the whole thing:

```{r frog-confint}
confint(frog_lm)
```

**Interpretation.** Each extra degree Celsius is associated with 2.33 more calls
per minute (95% CI: 1.60 to 3.06). We found this interval using a method that
captures the true population slope 95% of the time. The interval excludes zero,
which is the same conclusion the p-value gave.

# Inference for prediction

A point prediction at 21 degrees:

$$\hat{y} = -6.190332 + 2.330816(21) = 42.8$$

But a *prediction for the average frog* and a *prediction for one frog* carry
different amounts of uncertainty:

$$SE_{\hat{\mu}}=s\sqrt{\frac{1}{n} + \frac{(x^*-\bar{x})^2}{\sum(x-\bar{x})^2}}
\qquad
SE_{\hat{y}}=s\sqrt{1 + \frac{1}{n} + \frac{(x^*-\bar{x})^2}{\sum(x-\bar{x})^2}}$$

The only difference is the extra `1` under the second square root. That is the
variability of an individual frog around the line.

```{r frog-predict}
# the x value you want the prediction for
newdata <- data.frame(temp = 21)

# interval for a single new frog
predict(frog_lm, newdata, interval = "predict")

# interval for the mean response at temp = 21
predict(frog_lm, newdata, interval = "confidence")
```

```{r frog-intervals-plot, out.width="90%", fig.width=9, fig.height=3.5}
plot_a <- ggplot(frog_data, aes(x = temp, y = freq)) +
  geom_smooth(method = "lm", se = FALSE) +
  annotate("point", x = 21, y = 42.8, col = "orange", size = 3, pch = 4) +
  annotate("segment", x = 21, xend = 21, y = 40.38472, yend = 45.12887,
           col = "red", linewidth = 1) +
  geom_point() +
  theme_minimal(base_size = 13) +
  labs(x = "Temperature (Celsius)", y = "Frequency of mating calls",
       title = "95% Confidence Interval")

plot_b <- ggplot(frog_data, aes(x = temp, y = freq)) +
  geom_smooth(method = "lm", se = FALSE) +
  annotate("point", x = 21, y = 42.8, col = "orange", size = 3, pch = 4) +
  annotate("segment", x = 21, xend = 21, y = 36.37187, yend = 49.1473,
           col = "orange", linewidth = 1) +
  geom_point() +
  theme_minimal(base_size = 13) +
  labs(x = "Temperature (Celsius)", y = "Frequency of mating calls",
       title = "95% Prediction Interval")

plot_a + plot_b
```

The prediction interval is nearly three times wider. The confidence interval is
only uncertain about **where the line is**. The prediction interval is uncertain
about where the line is *and* about where one frog will land relative to it.

Notice that plenty of real frogs fall outside the narrow red interval. That is
correct. It was never meant to contain individuals.

# Practice problems

## Problem 1

Using `mana_lm`, test whether the population slope is zero. Pull the test
statistic out of `tidy()`, confirm it by hand from the estimate and its standard
error, and get the p-value from `pt()`. How many degrees of freedom does this
test have?

*Hint: `nrow(mana_data)` gives you n.*

## Problem 2

Build a 95% confidence interval for the manatee slope, first by hand with `qt()`
and then with `confint()`. Interpret it in a sentence a wildlife manager would
understand.

## Problem 3

Florida expects 900 thousand registered powerboats next year.

(a) Predict the number of manatee deaths.
(b) Give a 95% interval for the *average* number of deaths in years with 900
    thousand boats, and a 95% interval for the deaths in *one particular* such
    year.
(c) Which is wider, and why?
(d) Should you trust a prediction at 1,500 thousand boats? Why or why not?

## Problem 4

Check the assumptions for `mana_lm`. Is the inference you did in problems 1 to 3
justified?

# Answers

## Problem 1

```{r answer-1}
nrow(mana_data)          # n = 40, so df = n - 2 = 38
tidy(mana_lm)

# confirm the statistic by hand
b_hat  <- 0.1358071
se_b   <- 0.007640889
t_stat <- b_hat / se_b
t_stat

pt(q = t_stat, df = 38, lower.tail = FALSE) * 2
```

The test has $n - 2 = 38$ degrees of freedom. The slope sits about 17.8 standard
errors from zero, giving a p-value of about $5 \times 10^{-20}$. We reject $H_0$.
There is very strong evidence of an association between powerboat registrations
and manatee deaths.

## Problem 2

```{r answer-2}
t_star_mana <- qt(p = 0.975, df = 38)
t_star_mana

0.1358071 + c(-1, 1) * t_star_mana * 0.007640889

confint(mana_lm)
```

Each additional thousand registered powerboats is associated with about 0.136
more manatee deaths per year (95% CI: 0.120 to 0.151). Scaled up to something a
manager can act on, ten thousand extra boats is associated with roughly 1.2 to
1.5 extra manatee deaths per year.

## Problem 3

```{r answer-3}
new_boats <- data.frame(powerboats = 900)

# (a) and (b)
predict(mana_lm, new_boats, interval = "confidence")
predict(mana_lm, new_boats, interval = "predict")
```

(a) The point prediction is about 75.5 deaths.

(b) Confidence interval for the mean response: 72.0 to 79.0. Prediction interval
for a single year: 57.3 to 93.7.

(c) The prediction interval is much wider (about 36 deaths wide, versus 7). It
has to account for two things rather than one, the uncertainty about where the
true line sits *and* the year to year scatter of individual years around that
line.

(d) No. The data only cover 447 to 1027 thousand boats. At 1,500 you are
extrapolating well outside the observed range, where you have no evidence that
the relationship is still linear. Notice the $(x^*-\bar{x})^2$ term in the
standard error formula also grows, so the model itself widens the interval as you
move away from the centre of the data.

## Problem 4

```{r answer-4, out.width="85%"}
diag_plots(mana_lm, "Manatee model diagnostics")
```

Panel (c) shows no smile and no clear megaphone, so linearity and constant
variance are acceptable. The QQ plot in panel (b) is close to the line. Panel (d)
shows the residual box is far shorter than the deaths box, matching the r-squared
of 0.893.

The assumption to worry about here is **independence**, and no plot can check it.
These are 40 consecutive years from a single state, and both variables trend
upward over time, so successive years are unlikely to be truly independent. The
inference above is a reasonable first pass, but a careful analysis would account
for the time structure, and nothing here establishes that boats *cause* the
deaths.
