---
title: "L11:  The normal distribution"
---

<!-- libraries -->
```{r,include=FALSE,purl=FALSE}
library(knitr) # for include_graphics() 
library(dplyr)
library(ggplot2)
```

```{r load-libraries, message=F, warning=FALSE, echo=F}
#don't need to worry about knowing these libraries
library(jpeg)
library(grid)
library(patchwork)
library(latex2exp)
```


### The Normal Distribution N(0,1)

- Here is a Normal distribution with mean ($\mu$) = 0, standard deviation ($\sigma$) = 1 .

```{r load-libraries2, echo=F, out.width = "70%", warning=F, message=F}

#students, you don't need to know how to use stat_function()
p1 <- ggplot(data = data.frame(x = c(-3, 3)), aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = 0, sd = 1), col = "orange") +
  labs(y = "density") + 
  geom_text(aes(x = 0, y = 0.45), label = "mean = 0 \nsd = 1", check_overlap = T, col = "orange") +
  scale_x_continuous(limits = c(-5, 5)) + theme_minimal(base_size = 15)

p1
```

### The Normal Distribution N(0,1) and N(2,1)
- Let's add another Normal distribution, this one centered at 2, with the same
standard deviation

```{r, echo=F, out.width="80%"}
p1 <- p1 + stat_function(fun = dnorm, n = 101, args = list(mean = 2, sd = 1), col = "blue") +
  geom_text(aes(x = 2, y = 0.45), label = "mean = 2 \nsd = 1", check_overlap = T, col = "blue") 
p1
```

### The Normal Distribution N(O,1) and N(2,1) and N(-2,0.5)

- Let's add a third Normal distribution, this one centered at -2, with a standard
deviation of 0.5

```{r, echo=F, out.width="80%"}
p1 <- p1 + stat_function(fun = dnorm, n = 101, args = list(mean = -2, sd = 0.5), col = "forest green")  +
  geom_text(aes(x = -2, y = 0.85), label = "mean = -2 \nsd = 0.5", check_overlap = T, col = "forest green") 
p1
```

### The Normal Distribution

```{r, echo=F, out.width="80%"}
p1 <- p1 + stat_function(fun = dnorm, n = 101, args = list(mean = 1, sd = 1.5))  +
  geom_text(aes(x = 1, y = 0.3), label = "mean = 1 \nsd = 2.5", check_overlap = T) 
p1
```


```{r, echo=F, out.width="70%"}
#students, you don't need to know this code

p1 <- ggplot(data = data.frame(x = c(56, 73)), aes(x)) +
  stat_function(fun = dnorm, args = list(mean = 64.5, sd = 2.5)) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = 64.5, sd = 2.5), fill = "#ec7014", xlim = c(57, 72)) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = 64.5, sd = 2.5), fill = "#fe9929", xlim = c(59.5, 69.5), alpha = 0.8) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = 64.5, sd = 2.5), fill = "#fee391", xlim = c(62, 67), alpha = 0.8) +
  labs(y = "density", x = "") + 
  geom_text(aes(x = 64.5, y = 0.18), label = "mean = 64.5 \nsd = 2.5", check_overlap = T) +
  theme_minimal(base_size = 15) +
  geom_segment(aes(x = 64.5+2.5 , y = 0, xend = 64.5+2.5, yend = 0.1), lty = 1) +
  geom_segment(aes(x = 64.5-2.5 , y = 0, xend = 64.5-2.5, yend = 0.1), lty = 1) +
  geom_segment(aes(x = 64.5+2*2.5 , y = 0, xend = 64.5+2*2.5, yend = 0.1), lty = 2) +
  geom_segment(aes(x = 64.5-2*2.5 , y = 0, xend = 64.5-2*2.5, yend = 0.1), lty = 2) +
  geom_segment(aes(x = 64.5+3*2.5 , y = 0, xend = 64.5+3*2.5, yend = 0.1), lty = 3) +
  geom_segment(aes(x = 64.5-3*2.5 , y = 0, xend = 64.5-3*2.5, yend = 0.1), lty = 3) +
  scale_x_continuous(breaks = c(64.5-2*2.5, 64.5-2.5, 64.5, 64.5+2.5, 64.5+2*2.5), 
                     labels = c("59.5\n-2 SD", "62\n-1 SD", "64.5\nmean", "67\n+1 SD", "69.5\n+2 SD"),
                     limits = c(56, 73)) +
  geom_segment(aes(x = 64.5-2.5, xend = 64.5 + 2.5, y = 0.075, yend = 0.075)) +
  geom_label(aes(x = 64.5, y = 0.075), label = "68%", fill = "#fee391") +
  geom_segment(aes(x = 64.5-2*2.5, xend = 64.5 + 2*2.5, y = 0.05, yend = 0.05), lty = 2) +
  geom_label(aes(x = 64.5, y = 0.05), label = "95%", fill = "#fe9929") +
  geom_segment(aes(x = 64.5-3*2.5, xend = 64.5 + 3*2.5, y = 0.025, yend = 0.025), lty = 3) +
  geom_label(aes(x = 64.5, y = 0.025), label = "99.7%", fill = "#ec7014") 
  

p1
```

### Finding Normal probabilities

- A cumulative probability for a value x in a distribution is the proportion of 
observations in the distribution that lie at or below x.
- Here is the cumulative probability for x=1.2

```{r, out.width="70%", echo=F}
ggplot(data = data.frame(x = c(-4, 4)), aes(x)) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = 0, sd = 1), fill = "#ec7014", xlim = c(-4, 1.2)) +
  theme_minimal(base_size = 15) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1)) +
  geom_segment(x = 1.2, xend = 1.2, y = 0 , yend = 0.195) + 
  scale_x_continuous(breaks = -4:4) + 
  labs(y = "density")
```

### Finding Normal probabilities

- Recall that 100% of the sample space for the random variable x lies under the 
probability density function.
- What is the amount of the area that is below x = 1.2?
- To answer this question we use the `pnorm()` function:

```{r cumulative-prob-below}
pnorm(q = 1.2, mean = 0, sd = 1)
```

### Finding Normal probabilities

What if we wanted the reverse: P(x>1.2)?

```{r, out.width="70%", echo=F}
ggplot(data = data.frame(x = c(-4, 4)), aes(x)) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = 0, sd = 1), fill = "#ec7014", xlim = c(1.2, 4)) +
  theme_minimal(base_size = 15) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1)) +
  geom_segment(x = 1.2, xend = 1.2, y = 0 , yend = 0.195) + 
  scale_x_continuous(breaks = -4:4) + labs(y = "density")
```

```{r probability-above}
1 - pnorm(q = 1.2, mean = 0, sd = 1)
```

Alternatively: 

```{r alternative-prob-above}
 pnorm(q = 1.2, mean = 0, sd = 1, lower.tail = F)
```

So, 11.51% of the data is above x=1.2.

### Finding Normal probabilities

What if we wanted two "tail" probabilities?: $P(x< -1.2 \text{ or } x>1.2)$?

```{r, echo=F, out.width="70%"}
ggplot(data = data.frame(x = c(-4, 4)), aes(x)) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = 0, sd = 1), fill = "#ec7014", xlim = c(1.2, 4)) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = 0, sd = 1), fill = "#ec7014", xlim = c(-4, -1.2)) +
  theme_minimal(base_size = 15) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1)) +
  geom_segment(x = 1.2, xend = 1.2, y = 0 , yend = 0.195) + 
  geom_segment(x = -1.2, xend = -1.2, y = 0 , yend = 0.195) + 
  scale_x_continuous(breaks = -4:4) + labs(y = "density")
```

### Finding Normal probabilities

The trick: find one of the tails and then double the area because the distribution
is symmetric:

```{r}
pnorm(q = -1.2, mean = 0, sd = 1)*2
```

### Finding Normal probabilities

What if we wanted a range in the middle?: $P(-0.8 < x < 1.5)$?

```{r, echo=F, out.width="80%"}
ggplot(data = data.frame(x = c(-4, 4)), aes(x)) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = 0, sd = 1), fill = "#ec7014", xlim = c(-0.8, 1.5)) +
  theme_minimal(base_size = 15) +
  stat_function(fun = dnorm, args = list(mean = 0, sd = 1)) +
  geom_segment(x = 1.5, xend = 1.5, y = 0 , yend = 0.13) + 
  geom_segment(x = -0.8, xend = -0.8, y = 0 , yend = 0.29) + 
  scale_x_continuous(breaks = -4:4) + labs(y = "density")
```

### Finding Normal probabilities

```{r steps}
# step 1: calculate the probability *below* the upper bound (x=1.5)
pnorm(q = 1.5, mean = 0, sd = 1)
# step 2: calculate the probability *below* the lower bound (x = -0.8)
pnorm(q = -0.8, mean = 0, sd = 1)
# step 3: take the difference between these probabilities 
pnorm(q = 1.5, mean = 0, sd = 1) - pnorm(q = -0.8, mean = 0, sd = 1)
```

Thus, 72.13% of the data is in the range -0.8 < x < 1.5.

### Your turn 

To diagnose osteoporosis, bone mineral density is measured. The WHO criterion for 
osteoporosis is a BMD score below -2.5. Women in their 70s have a much lower BMD
than younger women. Their BMD ~ N(-2, 1). What proportion of these women have a
BMD below the WHO cutoff?

```{r osteoporosis, echo=FALSE, include=FALSE}
#to fill in during class

pnorm(q=-2.5, mean=-2, sd= 1 )
```


### Finding Normal percentiles

Example: The hatching weights of commercial chickens can be modeled accurately
using a Normal distribution with mean $\mu=45$ grams and standard deviation 
$\sigma=4$ grams. What is the third quartile of the distribution of hatching 
weights?

```{r, echo=F, out.width="70%"}
ggplot(data = data.frame(x = c(29, 61)), aes(x)) +
  geom_area(stat = "function", fun = dnorm, args = list(mean = 45, sd = 4), fill = "#ec7014", xlim = c(29, 47.7)) +
  theme_minimal(base_size = 15) +
  stat_function(fun = dnorm, args = list(mean = 45, sd = 4)) +
  geom_segment(x = 47.69796, xend = 47.69796, y = 0 , yend = 0.08) + 
  labs(y = "density") +
  geom_text(aes(x = 44, y = 0.05), check_overlap = T, label = "75%", col = "white", size = 8) 
```

### Finding Normal percentiles using the `qnorm()` function

Example: The hatching weights of commercial chickens can be modeled accurately
using a Normal distribution with mean $\mu=45$ grams and standard deviation 
$\sigma=4$ grams. What is the third quartile of the distribution of hatching 
weights?

```{r}
qnorm(p = 0.75, mean = 45, sd = 4)
```

Thus, 75% of the data is below 47.7 for this distribution.


## The standard normal and Z scores

### The standard Normal distribution
- The standard Normal distribution N(0,1) has $\mu = 0$ and $\sigma=1$.
- $X \sim N(0,1)$, implies that the random variable X is Normally distributed.

```{r, echo=F, out.width="75%"}
p1 <- ggplot(data = data.frame(x = c(-3, 3)), aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = 0, sd = 1), col = "orange") +
  labs(y = "density", x = "") + 
  geom_text(aes(x = 0, y = 0.45), label = "mean = 0 \nsd = 1", check_overlap = T, col = "orange") +
  theme_minimal(base_size = 15) +
  geom_segment(aes(x = 0 , y = 0, xend = 0, yend = 0.4)) +
  geom_segment(aes(x = 1 , y = 0, xend = 1, yend = 0.25), lty = 2) +
  geom_segment(aes(x = -1 , y = 0, xend = -1, yend = 0.25), lty = 2) +
  geom_segment(aes(x = 2 , y = 0, xend = 2, yend = 0.05), lty = 3) +
  geom_segment(aes(x = -2 , y = 0, xend = -2, yend = 0.05), lty = 3) +
  scale_x_continuous(breaks = -3:3, 
                     labels = c(-3, "-2\n-2 SD", "-1\n-1 SD", "0\nmean", "1\n+1 SD", "2\n+2 SD", 3))
p1
```

## Simulating Normally distributed data in R

### Simulating Normally distributed data in R

Suppose that we measured 1000 heights for young women:

```{r simulate-height-data}
#students, rnorm() is important to know!
heights.women <- rnorm(n = 1000, mean = 64.5, sd = 2.5)
heights.women <- data.frame(heights.women)
```

### Simulating Normally distributed data in R
We can plot the histogram of the heights, and see that they roughly follow from
a Normal distribution:

```{r plot heights, out.width="80%", echo=F, warning=F, message=F}
hist <- ggplot(heights.women, aes(x = heights.women)) + geom_histogram(col = "white") + theme_minimal(base_size = 15) 

dens <- ggplot(heights.women, aes(x = heights.women)) + geom_density() + stat_function(fun = dnorm, args = c(mean = 64.5, sd = 2.5), col = "forest green") + theme_minimal(base_size = 15)

hist + dens + plot_layout()
```

### Standardizing Normally distributed data in R

To standard these data, we can apply the formula to compute the z-value:

```{r compute-z-value}
heights.women <- heights.women %>% mutate(mean = mean(heights.women), 
                                          sd = sd(heights.women), 
                                          z = (heights.women - mean)/sd)

head(heights.women)
```

What would the distribution of the standardized heights look like?

### Standardizing Normally distributed data in R

```{r plot-z-scores, warning=F, message=F, out.width="80%", echo=F, }
hist_z <- ggplot(heights.women, aes(x = z)) + geom_histogram(col = "white") + 
  theme_minimal(base_size = 15) 

dens_z <- ggplot(heights.women, aes(x = z)) + geom_density() + 
  stat_function(fun = dnorm, args = c(mean = 0, sd = 1), col = "forest green") + 
  theme_minimal(base_size = 15)

hist_z + dens_z + plot_layout()
```

### Are these data Normally distributed?

- The data is unimodal and symmetric, but is its distribution Normal?

```{r, echo=F, message=FALSE, out.width="75%"}
library(ggplot2)
example_data <- data.frame(y = rt(200, df = 5))

ggplot(example_data, aes(y)) + geom_histogram(col = "white") +
  theme_minimal(base_size = 15)
```

### Making a QQ plot step by step

1. First, arrange the variable in ascending order. Calculate the percentile for each measurement.
For example if you had ten measurements in ascending order, the first measurement is at the 
10th percentile because 10% of the data is at or below its value. The second measurement
is at the 20% because 20% of the data is at or below its value, and so forth.
2. Then, for each of the percentiles you calculated, use that percentile to calculate
the corresponding x-value of the Normal distribution that occurs at that percentile.
For example, at x = -1 at the 16th percentile of the N(0, 1) distribution.
3. Make a scatter plot of the calculated x-values on the x-axis and the original
variable values on the y-axis.
4. The closer the data is to a straight line, the more closely it approximates
a Normal distribution.

### Making a QQ plot step by step

```{r qqplot-calcs, message=FALSE}
#1. calculate the percentile :
example_data <- example_data %>%  arrange(y) %>% 
  mutate(quantile = row_number()/n())
# 2. then calculate the x-value at each percentile from the previous step
# 3. this x-value can be called a z-score because it is from the standard Normal distribution
example_data <- example_data %>% 
  mutate(z_score = qnorm(quantile, mean = 0, sd = 1))
head(example_data)
```

### Look at the QQ plot for these data

- Notice that the data overlays the 45 degree line in the middle but not in the 
tails of the distribution. This sort of pattern shows that these data are "wider"
(have larger standard deviation) that a Normally distributed variable.

```{r qqplots, out.width="75%", fig.align=T, echo=FALSE}
ggplot(example_data, aes(x = z_score, y = y)) + 
  geom_point() + 
  geom_abline(intercept = 0, slope = 1) + 
  theme_minimal(base_size = 15)
```

### Easy way to make a qqplot() where R does all the calculating for you

```{r, out.width="75%", fig.align=T}
ggplot(example_data, aes(sample = y)) + stat_qq() + stat_qq_line() + 
  theme_minimal(base_size = 15)
```

### Another example:  Seed Data
```{r, echo=FALSE}
library(readr)
seed_data <- read_csv("Ch04_seed-data")

head(seed_data)
```

### Another example
Check out its distribution. It definitely does not look normal:

```{r, out.width="70%", fig.align='center', echo=FALSE, message=FALSE}
ggplot(seed_data, aes(x = seed_count)) + 
  geom_histogram(col = "white") + 
  theme_minimal(base_size = 15)
```

### Another example

And look at its QQ plot. Does the data appear to follow a Normal distribution? 

```{r, echo=FALSE}
ggplot(seed_data, aes(sample = seed_count)) + 
  stat_qq() + stat_qq_line() + 
  theme_minimal(base_size = 15)
```

### Another example (logged)

You might remember that we took the log of seed_count before we used it in regression. The log values look like this:

```{r, out.width="70%", echo=FALSE}
ggplot(seed_data, aes(x = log(seed_count))) + 
  geom_histogram(col = "white", binwidth = 2) + 
  theme_minimal(base_size = 15)
```

### Another example (logged)

How does the QQ plot look for the logged variable?

```{r, echo=FALSE}
ggplot(seed_data, aes(sample = log(seed_count))) + 
  stat_qq() + stat_qq_line() + 
  theme_minimal(base_size = 15) + scale_y_continuous(limits = c(0, 15)) 
```


