---
title: "L23: Inference for a population proportion"
output: html_document
---

```{r load-libraries, include=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(tibble)
library(ggplot2)
knitr::opts_chunk$set(fig.align = "center", message = FALSE, warning = FALSE)
options(width = 90)
```

Every calculation from L23 in one runnable place. Up to now every outcome has been
continuous, so we summarized it with a mean and used a $t$ distribution. Now the
outcome is categorical with two levels, so we summarize it with a **proportion**
and go back to the **Normal** distribution.

There is no external data file. Every number below is typed in from the slides.
Try the practice problems at the bottom before checking the answers.

## The four confidence interval methods

| Method | Formula / function | Do it by hand? |
|--------|--------------------|----------------|
| Large sample | $\hat{p} \pm z^*\sqrt{\hat{p}(1-\hat{p})/n}$ | yes |
| Plus four | $\tilde{p} \pm z^*\sqrt{\tilde{p}(1-\tilde{p})/(n+4)}$, with $\tilde{p} = \frac{x+2}{n+4}$ | yes |
| Wilson score (with continuity correction) | `prop.test(x, n)` | no, R does it |
| Clopper Pearson ("exact") | `binom.test(x, n)` | no, R does it |

Two things to keep straight:

- We use $z^*$, not $t^*$. The standard error of $\hat{p}$ is fully determined by
  $p$ and $n$, so there is no second parameter to estimate.
- **Check the condition first.** The Normal approximation needs at least 10
  expected successes *and* at least 10 expected failures.

## The plus four calculation by hand

A random sample of 75 SARS patients, of which 64 developed recurrent fever.

```{r sars-plus-four}
x <- 64
n <- 75

p.hat <- x / n                            # ordinary estimate
p.tilde <- (x + 2) / (n + 4)              # plus four estimate
se <- sqrt(p.tilde * (1 - p.tilde) / (n + 4))

c(p.hat, p.tilde, se)

p.tilde + c(-1, 1) * 1.96 * se            # 95% CI
```

So the plus four 95% CI is 75.4% to 91.7%. The most common mistake on this
calculation is forgetting the `1.96`, which gives a far too narrow interval.

# Example 1: 100 deaths out of 500 hip fractures

First check the condition. $n\hat{p} = 100$ and $n(1-\hat{p}) = 400$, both well
above 10, so we expect all four methods to agree.

## Large sample method

```{r ex1-large-sample}
p.hat <- 100/500                          # estimate proportion
se <- sqrt(p.hat*(1-p.hat)/500)           # standard error
c(p.hat - 1.96*se, p.hat + 1.96*se)       # CI
```

16.5% to 23.5%. Note this is pure arithmetic, no statistical function is called.

## Plus four method

```{r ex1-plus-four}
p.tilde <- (100 + 2)/(500 + 4)
se <- sqrt(p.tilde*(1-p.tilde)/504)       # standard error
c(p.tilde - 1.96*se, p.tilde + 1.96*se)   # CI
```

16.7% to 23.7%. With $n = 500$, adding four imaginary observations barely moves
the answer.

## Wilson score method

```{r ex1-wilson}
prop.test(x = 100, n = 500, conf.level = 0.95)
```

Read the `95 percent confidence interval` line: 16.6% to 23.8%. Ignore the rest.
`prop.test` also ran a hypothesis test against a null of $p_0 = 0.5$ that you did
not ask for. To pull out just the interval:

```{r ex1-wilson-ci}
prop.test(x = 100, n = 500)$conf.int
```

## Clopper Pearson exact method

```{r ex1-exact}
binom.test(x = 100, n = 500, conf.level = 0.95)$conf.int
```

16.6% to 23.8%. This method makes no Normal approximation at all, and it still
lands within a tenth of a percentage point of the large sample answer.

## Comparing the four

```{r ex1-plot, out.width="75%", fig.width=7, fig.height=3.4}
elderly_CIs <- tibble(
  method   = c("Large sample", "Plus 4", "Wilson", "Exact"),
  lower    = c(16.5, 16.7, 16.6, 16.6),
  upper    = c(23.5, 23.7, 23.8, 23.8),
  estimate = rep(20, 4))

elderly_CIs$method <- factor(elderly_CIs$method,
  levels = c("Large sample", "Plus 4", "Wilson", "Exact"))

ggplot(data = elderly_CIs, aes(x = method, y = estimate)) +
  geom_point() +
  geom_segment(aes(xend = method, y = lower, yend = upper)) +
  geom_hline(aes(yintercept = 0), lty = 2) +
  scale_y_continuous(limits = c(0, 26)) +
  labs(y = "Estimate and 95% CI (%)", x = "") +
  theme_minimal(base_size = 13)
```

The picture is boring, and that is the point. When the conditions are comfortably
satisfied, the choice of method does not matter.

Note that only the large sample interval is **symmetric** about $\hat{p} = 20\%$.
That is fine. Nothing requires a confidence interval to be symmetric, and near
the boundaries of 0 and 1 the honest interval leans away from the wall.

# Example 2: 2 deaths out of 100 falls

Check the condition first. $n\hat{p} = 2$, far below 10. The Normal approximation
is **not** trustworthy here.

## Large sample and plus four

```{r ex2-hand}
p.hat <- 2/100                            # estimate proportion
se <- sqrt(p.hat*(1-p.hat)/100)           # standard error
c(p.hat - 1.96*se, p.hat + 1.96*se)       # CI

p.tilde <- (2 + 2)/(100 + 4)
se <- sqrt(p.tilde*(1-p.tilde)/104)       # standard error
c(p.tilde - 1.96*se, p.tilde + 1.96*se)   # CI
```

Look at the large sample lower bound: **negative**. A proportion cannot be
negative. The method has broken, and it broke silently.

## Exact and Wilson

```{r ex2-r}
binom.test(x = 2, n = 100, conf.level = 0.95)$conf.int
prop.test(x = 2, n = 100, conf.level = 0.95)$conf.int
```

## All four side by side

| Method | 95% CI | R function |
|--------|:------:|------------|
| Large sample | -0.74% to 4.74% | by hand |
| Plus four | 0.15% to 7.54% | by hand |
| Wilson score | 0.35% to 7.74% | `prop.test` |
| Exact | 0.24% to 7.04% | `binom.test` |

```{r ex2-plot, out.width="75%", fig.width=7, fig.height=3.4}
elderly_CIs <- tibble(
  method   = c("Large sample", "Plus 4", "Wilson", "Exact"),
  lower    = c(-0.74, 0.15, 0.35, 0.24),
  upper    = c(4.74, 7.54, 7.74, 7.04),
  estimate = rep(2, 4))

elderly_CIs$method <- factor(elderly_CIs$method,
  levels = c("Large sample", "Plus 4", "Wilson", "Exact"))

ggplot(data = elderly_CIs, aes(x = method, y = estimate)) +
  geom_point() +
  geom_segment(aes(xend = method, y = lower, yend = upper)) +
  geom_hline(aes(yintercept = 0), lty = 2) +
  labs(y = "Estimate and 95% CI (%)", x = "") +
  theme_minimal(base_size = 13)
```

Three findings:

- The intervals genuinely disagree. In Example 1 the choice of method was
  irrelevant; here it changes your answer.
- The large sample lower bound is nonsensical.
- It broke because $n\hat{p} = 2 < 10$. The condition is not a formality.

# Hypothesis test for a proportion

The null supplies a specific value $p_0$ from **outside** the data:
$H_0: p = p_0$.

$$z = \frac{\hat{p}-p_0}{\sqrt{\frac{p_0(1-p_0)}{n}}}$$

The standard error uses $p_0$, **not** $\hat{p}$. A test asks what the world looks
like *if the null is true*, and the null tells us the proportion, so there is
nothing left to estimate. A confidence interval has no null to lean on, so it
falls back on $\hat{p}$.

Condition: $np_0 \geq 10$ and $n(1-p_0) \geq 10$. Again $p_0$, not $\hat{p}$.

## Worked example

200 patients, 33 with reduced or no side-effects, against a historical rate of
10%. $H_0: p = 0.10$ vs. $H_a: p > 0.10$.

```{r ht-by-hand}
p.hat <- 33/200
p.0 <- 0.10
n <- 200

n * p.0            # expected successes under H0, need >= 10
n * (1 - p.0)      # expected failures under H0, need >= 10

se <- sqrt(p.0 * (1 - p.0) / n)            # note p.0, not p.hat
z <- (p.hat - p.0) / se
c(p.hat, se, z)

pnorm(q = z, lower.tail = FALSE)           # one-sided p-value
```

The p-value is about 0.0011. If the true rate really were 10%, a sample proportion
of 16.5% or higher out of 200 patients would happen about once in a thousand
times. That is strong evidence against the 10% claim.

## The same test in R

```{r ht-prop-test}
prop.test(x = 33, n = 200, p = 0.10,
          alternative = "greater", correct = FALSE)
```

Same p-value, 0.001092. Two things to notice:

- Setting `p = 0.10` is essential. Leave it out and R silently tests $p_0 = 0.5$.
- R prints `X-squared` rather than $z$, because a squared standard Normal is a
  chi-square with 1 degree of freedom. Indeed $3.064^2 = 9.389$.

`correct = FALSE` turns off the continuity correction so the answer matches the
hand calculation. In real work, leave the correction on.

# Sample size for a proportion

$$n = \left(\frac{z^*}{m}\right)^2 p^*(1-p^*)$$

where $m$ is the desired **margin of error** (the *half* width of the interval)
and $p^*$ is your best guess at the proportion. With no guess, use $p^* = 0.5$,
which maximizes $p(1-p)$ and so gives the most conservative (largest) $n$.

**Always round up.**

```{r sample-size}
n_for_prop <- function(m, p.star, conf = 0.95) {
  z.star <- qnorm(1 - (1 - conf)/2)
  ceiling((z.star / m)^2 * p.star * (1 - p.star))
}

n_for_prop(m = 0.04, p.star = 0.5)     # no prior information
n_for_prop(m = 0.04, p.star = 0.25)    # prior study suggested 25%
n_for_prop(m = 0.03, p.star = 0.5)     # CI *width* of 6 points
```

Halving the margin of error roughly quadruples the sample size, because of the
square in the formula.

## The Twitter bot question

Twitter sampled 9,000 accounts per quarter out of 229 million users, and reported
fewer than 5% bots. How large a sample is needed for a margin of error of 0.5%?

```{r twitter}
n_for_prop(m = 0.005, p.star = 0.04)
```

5,901. Twitter sampled 9,000, so their sample was comfortably large enough.

Notice that **229 million never appears in the formula**. The precision of an
estimate depends on how many things you looked at, not on how many you did not.

---

# Practice problems

Try these before looking at the answers.

**Problem 1.** In a random sample of 250 adults, 40 reported a particular
symptom. (a) Compute the large sample 95% CI by hand. (b) Compute the plus four
interval. (c) Get the Wilson and exact intervals from R. (d) Should the four
agree here? Explain using the condition.

**Problem 2.** In a small pilot study, 3 of 40 participants had an adverse event.
(a) Compute the large sample 95% CI. (b) What is wrong with it? (c) Which
interval would you report instead, and why?

**Problem 3.** A manufacturer claims that at most 2% of its devices are
defective. In a random sample of 400 devices, 14 are defective. (a) State the
hypotheses. (b) Check the condition. (c) Compute $z$ and the p-value by hand.
(d) Confirm with `prop.test`. (e) State the conclusion in the language of the
problem.

**Problem 4.** You want to estimate a vaccination rate with a 95% confidence
interval no wider than 4 percentage points. (a) What is $m$? (b) How many people
do you need with no prior information? (c) A previous survey suggests the rate is
about 80%. How many now? (d) How much did the previous survey save you?

---

# Answers

## Problem 1

```{r answer-1}
x <- 40; n <- 250

# (a) large sample
p.hat <- x/n
se <- sqrt(p.hat*(1-p.hat)/n)
p.hat + c(-1, 1) * 1.96 * se

# (b) plus four
p.tilde <- (x + 2)/(n + 4)
se4 <- sqrt(p.tilde*(1-p.tilde)/(n + 4))
p.tilde + c(-1, 1) * 1.96 * se4

# (c) Wilson and exact
prop.test(x, n)$conf.int
binom.test(x, n)$conf.int

# (d) condition
c(n*p.hat, n*(1-p.hat))
```

(d) Yes, they should agree. $n\hat{p} = 40$ and $n(1-\hat{p}) = 210$, both well
above 10, so the Normal approximation is sound. The four intervals all sit in the
region 11.5% to 21.3%, and no two of them differ by as much as a percentage point
at either end. Any of the four supports the same conclusion.

## Problem 2

```{r answer-2}
x <- 3; n <- 40

# (a)
p.hat <- x/n
se <- sqrt(p.hat*(1-p.hat)/n)
p.hat + c(-1, 1) * 1.96 * se

n * p.hat   # expected successes

# (c)
p.tilde <- (x + 2)/(n + 4)
se4 <- sqrt(p.tilde*(1-p.tilde)/(n + 4))
p.tilde + c(-1, 1) * 1.96 * se4

binom.test(x, n)$conf.int
```

(b) The lower bound is negative, which is impossible for a proportion. The
approximation failed because $n\hat{p} = 3$, well below the required 10.

(c) Report the plus four interval if working by hand (about 2.0% to 20.7%), or
the exact interval from `binom.test` (about 1.6% to 20.4%). Both stay positive,
and both are wider than the large sample interval, which is the honest answer
given only 40 observations.

## Problem 3

```{r answer-3}
x <- 14; n <- 400; p.0 <- 0.02

# (b) condition, using p.0 not p.hat
c(n*p.0, n*(1-p.0))

# (c)
p.hat <- x/n
se <- sqrt(p.0*(1-p.0)/n)
z <- (p.hat - p.0)/se
c(p.hat, se, z)
pnorm(z, lower.tail = FALSE)

# (d)
prop.test(x = x, n = n, p = p.0, alternative = "greater", correct = FALSE)
```

(a) $H_0: p = 0.02$ vs. $H_a: p > 0.02$. The claim is "at most 2%", so the
scientifically interesting departure is a *higher* defect rate, which makes this
one-sided. Choose the direction before seeing the data.

(b) $np_0 = 8$, which is just below 10, while $n(1-p_0) = 392$ is fine. The
condition is marginally violated, so treat the p-value as approximate. `binom.test`
would be the safer choice here.

(c) $\hat{p} = 3.5\%$, $SE = 0.007$, $z = 2.14$, one-sided p-value about 0.016.

(e) If the true defect rate were 2%, we would see 14 or more defectives in 400
devices about 1.6% of the time. That is reasonably strong evidence against the
manufacturer's claim, and in favour of a defect rate above 2%. Given the marginal
condition, we would want to confirm with the exact test before acting on it.

## Problem 4

```{r answer-4}
# (a) width 4 points => m = 0.02
n_for_prop(m = 0.02, p.star = 0.5)     # (b)
n_for_prop(m = 0.02, p.star = 0.8)     # (c)

n_for_prop(m = 0.02, p.star = 0.5) - n_for_prop(m = 0.02, p.star = 0.8)
```

(a) $m = 0.02$. The margin of error is the *half* width, so a width of 4
percentage points means $m$ is 2 percentage points. This is the single most common
error on sample size questions.

(b) 2401 people.

(c) 1537 people, because $0.8 \times 0.2 = 0.16$ is well below the worst case
$0.5 \times 0.5 = 0.25$.

(d) 864 fewer people, about a 36% reduction in the size of the survey, for the
price of a literature search. This is why pilot studies and prior surveys are
worth reading before you design a study.
