---
title: "L21: Non-parametric testing"
output: html_document
---

```{r load-libraries, include=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
knitr::opts_chunk$set(fig.align = "center", message = FALSE, warning = FALSE)
```

Every calculation from L21 in one runnable place. This lecture gives you a
rank-based alternative for each of the tests from L18 to L20, for when the
normality assumption is not something you want to lean on.

The only external file you need is `nhanes.csv`, which should sit in the same
folder as this file. Everything else is created in the code below. Try the
practice problems at the bottom before checking the answers.

## The idea

A parametric test uses the actual values, and assumes something about the shape
of the distribution they came from. A non-parametric test throws away the values
and keeps only their **rank order**, so it does not need that assumption.

The price is power. If the data really are normal, the rank test is the weaker
choice.

| Samples | Parametric | Non-parametric |
|---------|------------|----------------|
| Two independent samples | two sample t-test | Wilcoxon rank-sum |
| Two paired samples | paired t-test | Wilcoxon signed-rank |
| Three or more samples | ANOVA | Kruskal-Wallis |

## Wilcoxon rank-sum: the small example by hand

Group 1 is 4, 3, 5, 2, 6 and group 2 is 6, 5, 7, 4, 8. Pool all ten values and
rank them. Tied values share the average of the ranks they would have taken.

```{r rank-sum-by-hand}
g1 <- c(4, 3, 5, 2, 6)
g2 <- c(6, 5, 7, 4, 8)

all_ranks <- rank(c(g1, g2))
data.frame(value = c(g1, g2),
           group = c(rep("g1", 5), rep("g2", 5)),
           rank  = all_ranks)

W1 <- sum(all_ranks[1:5])
W2 <- sum(all_ranks[6:10])
c(sum_group1 = W1, sum_group2 = W2)
```

W is the **smaller** of the two sums. Then

$$\mu_w=\frac{n_s(n_s+n_l+1)}{2} \qquad \sigma_{w}=\sqrt{\frac{n_s n_l(n_s+n_l+1)}{12}} \qquad Z_{w}=\frac{W-\mu_w}{\sigma_{w}}$$

```{r rank-sum-z}
W  <- min(W1, W2)
ns <- 5
nl <- 5

mu_w    <- ns * (ns + nl + 1) / 2
sigma_w <- sqrt(ns * nl * (ns + nl + 1) / 12)
Z_w     <- (W - mu_w) / sigma_w

c(W = W, mu_w = mu_w, sigma_w = round(sigma_w, 3), Z_w = round(Z_w, 3))

# two sided p-value
2 * pnorm(-abs(Z_w))
```

A p-value of 0.09 means we do not reject the null at alpha = 0.05.

## Wilcoxon rank-sum in R

Unpaired is the **default**, so you do not need `paired = FALSE`. In current
versions of R the formula version will throw an error if you pass `paired` to it.

```{r rank-sum-syntax, eval=FALSE}
wilcox.test(group1, group2)                  # two vectors
wilcox.test(outcome ~ group, data = mydata)  # formula
```

## Rank-sum example: phenylketonuria

Normalized mental age scores for children with phenylketonuria, split by whether
their exposure was below or at least 10.0 mg/dl.

```{r pku-data}
nMA <- c(34.5, 37.5, 39.5, 40, 45.5, 47, 47, 47.5, 48.7,
         49, 51, 51, 52, 53, 54, 54, 55, 56.5, 57, 58.5,
         58.5, 28, 35, 37, 37, 43.5, 44, 45.5, 46, 48,
         48.3, 48.7, 51, 52, 53, 53, 54, 54, 55)
Group <- c(rep("low", 21), rep("high", 18))
pku <- data.frame(Group, nMA)

group_by(pku, Group) %>%
  summarise(count = n(),
            median = median(nMA),
            IQR = IQR(nMA))
```

```{r pku-density}
ggplot(pku, aes(x = nMA)) +
  geom_density(aes(fill = Group), alpha = 0.5) +
  theme_minimal(base_size = 14)
```

Neither group looks normal, and the low group is clearly bimodal, so the rank
test is the safer choice here.

```{r pku-test}
wilcox.test(nMA ~ Group, data = pku)
```

## Rank-sum versus t: NHANES heights

```{r nhanes-read}
nhanes <- read.csv("nhanes.csv")
dim(nhanes)

ht_m <- nhanes$bmxht[nhanes$gender == "Male"]
ht_f <- nhanes$bmxht[nhanes$gender == "Female"]
```

```{r nhanes-plots}
boxplot(bmxht ~ gender, data = nhanes, main = "Height in cm by gender")

ggplot(nhanes, aes(x = bmxht)) +
  geom_density(aes(fill = gender), alpha = 0.3) +
  theme_minimal(base_size = 14)
```

```{r nhanes-tests}
t.test(ht_m, ht_f, paired = FALSE)

wilcox.test(ht_m, ht_f, paired = FALSE)
```

With n over 2000 the two approaches agree completely. Both p-values are below
2.2e-16. **The tests disagree when the sample is small and the shape is bad, not
when the sample is large.**

## Wilcoxon signed-rank: the pre and post example by hand

Thirteen students take a test twice, before and after a review session.

```{r signed-rank-data}
test1 <- c(65, 87, 77, 90, 70, 84, 92, 83, 85, 91, 68, 72, 81)
test2 <- c(77, 100, 75, 89, 80, 81, 91, 96, 84, 89, 88, 100, 81)
Change <- test2 - test1
Change
```

Step 1 is the differences, above. Step 2 drops any pair whose difference is 0,
then ranks what is left by **absolute value**. Step 3 attaches the sign. Step 4
sums the positive and the negative ranks.

```{r signed-rank-table}
nonzero <- Change[Change != 0]
sr <- data.frame(difference = nonzero,
                 abs_diff   = abs(nonzero),
                 sign       = ifelse(nonzero > 0, "+", "-"),
                 rank       = rank(abs(nonzero)))
sr[order(sr$abs_diff), ]

sum_neg <- sum(sr$rank[sr$difference < 0])
sum_pos <- sum(sr$rank[sr$difference > 0])
c(sum_negative = sum_neg, sum_positive = sum_pos)
```

Note the 0 difference gets dropped, so n is 12 and not 13.

$$\mu_{T}=\frac{n(n+1)}{4} \qquad \sigma_{T}=\sqrt{\frac{n(n+1)(2n+1)}{24}} \qquad Z_{T}=\frac{T-\mu_{T}}{\sigma_{T}}$$

```{r signed-rank-z}
n <- length(nonzero)
Tstat <- min(sum_neg, sum_pos)

mu_T    <- n * (n + 1) / 4
sigma_T <- sqrt(n * (n + 1) * (2 * n + 1) / 24)
Z_T     <- (Tstat - mu_T) / sigma_T

c(n = n, T = Tstat, mu_T = mu_T, sigma_T = round(sigma_T, 3), Z_T = round(Z_T, 3))

2 * pnorm(-abs(Z_T))
```

## Wilcoxon signed-rank in R

```{r signed-rank-test}
wilcox.test(test1, test2, paired = TRUE, correct = FALSE)
```

R gives 0.157 and our hand calculation gave 0.158. The gap is the rounding in our
Z. The formula version needs `Pair()`, because it does not take a `paired`
argument.

```{r signed-rank-pair}
prepost <- data.frame(test1, test2)
wilcox.test(Pair(test1, test2) ~ 1, data = prepost, correct = FALSE)
```

Compare to the paired t-test.

```{r signed-rank-vs-t}
t.test(test1, test2, paired = TRUE)

hist(Change, main = "Change from time 1 to time 2")
```

Here the two tests **disagree**. The paired t gives 0.036 and rejects, the
signed-rank gives 0.157 and does not. With n = 13 and a lumpy distribution of
changes, that disagreement is the whole point of the lecture.

## Kruskal-Wallis

The non-parametric alternative to ANOVA. It compares medians and tells you that
at least one group differs, but not which one.

```{r kruskal}
treatment <- c(rep("Group1", 4), rep("Group2", 5),
               rep("Group3", 6), rep("Group4", 7))
outcome <- c(30, 46, 99, 95,               # group 1
             3, 14, 16, 41, 47,            # group 2
             5, 4, 4, 4, 10, 9,            # group 3
             24, 30, 43, 51, 62, 32, 96)   # group 4
KW_data <- data.frame(treatment, outcome)

kruskal.test(outcome ~ treatment, data = KW_data)
```

Try tinkering with this. Add an outlier to one group, or shift a whole group up,
and watch how much less the Kruskal-Wallis result moves than the ANOVA below.

```{r kruskal-vs-anova}
summary(aov(outcome ~ treatment, data = KW_data))
```

# Practice problems

## Problem 1

In which of these scenarios would we want to use a non-parametric test?

1. Jessica collects data on student study habits, but her sample is too small to
lean on the central limit theorem.
2. Chandler compares midterm scores for students who did and did not take a
practice midterm. Both distributions are approximately normal.
3. Mitchell and Antonia relate sleep hours to PH142 performance. Sleep hours are
left skewed with outliers on both sides, and they have only 20 students.
4. Alma and Yunzhe find the distribution of homework scores is bimodal.

*Hint: ask what the parametric test needs, and whether this scenario can supply it.*

## Problem 2

A study by Jalkanen et al. measured anti-S1 IgG concentration in one group of
volunteers before vaccination and a **different** group after vaccination.

| 0 days | 42 days |
|--------|---------|
| 30     | 111     |
| 26     | 112     |
| 23     | 107     |
| 21     | 111     |
| 19     | 127     |
| 17     | 108     |
| 25     | 115     |

Which test applies? Fill in the ranks, get the two rank sums, and compute Z by
hand. Then check yourself against `wilcox.test()`.

*Hint: the two groups are different people, so the samples are independent.*

## Problem 3

Go back to the pre and post test data. Suppose the last student had scored 81 at
time 1 and 95 at time 2, instead of 81 and 81. Redo the signed-rank test. Does the
conclusion change, and why does n change?

*Hint: nothing gets dropped now.*

# Answers

## Problem 1

**Scenarios 1, 3, and 4.**

1. Yes. With a small sample we cannot appeal to the CLT for normality of the
sampling distribution, and rank-based tests do not need it.
2. No. Both groups are approximately normal, so the parametric two sample t-test
is appropriate and has more power.
3. Yes. Skew plus outliers plus n = 20 is exactly the case rank-based tests were
built for.
4. Yes. A bimodal distribution is nowhere near normal, and the mean is a poor
summary of it.

## Problem 2

Independent samples with a continuous outcome, so this is the **Wilcoxon
rank-sum** test.

```{r answer-2}
d0  <- c(30, 26, 23, 21, 19, 17, 25)
d42 <- c(111, 112, 107, 111, 127, 108, 115)

r <- rank(c(d0, d42))
data.frame(value = c(d0, d42),
           group = c(rep("0 days", 7), rep("42 days", 7)),
           rank  = r)

W_0  <- sum(r[1:7])
W_42 <- sum(r[8:14])
c(sum_0days = W_0, sum_42days = W_42, total = W_0 + W_42)
```

The two 111 values tie for ranks 10 and 11, so both get 10.5. As a check, all the
ranks must sum to 14 times 15 over 2, which is 105.

```{r answer-2-z}
W  <- min(W_0, W_42)
ns <- 7
nl <- 7

mu_w    <- ns * (ns + nl + 1) / 2
sigma_w <- sqrt(ns * nl * (ns + nl + 1) / 12)
Z_w     <- (W - mu_w) / sigma_w

c(W = W, mu_w = mu_w, sigma_w = round(sigma_w, 3), Z_w = round(Z_w, 3))
2 * pnorm(-abs(Z_w))

wilcox.test(d0, d42)
```

We reject the null. IgG concentrations differ before and after vaccination.

## Problem 3

```{r answer-3}
test2_alt <- test2
test2_alt[13] <- 95        # was 81, tied with time 1

wilcox.test(test1, test2_alt, paired = TRUE, correct = FALSE)
```

Now no pair has a difference of 0, so nothing is dropped and n is 13 rather than
12. The extra positive rank pushes the positive sum further above the negative
sum, while T stays at the negative sum of 21. Since $\mu_T$ grows with n, Z moves
from -1.41 to -1.71 and the p-value falls from 0.157 to 0.086.

The conclusion does **not** change. We still fail to reject at alpha = 0.05. But
notice how far one student moved the p-value, which is what small samples do.
