---
title: "L24: Comparing two proportions"
output: html_document
---

```{r load-libraries, include=FALSE, message=FALSE, warning=FALSE}
library(tibble)
library(ggplot2)
knitr::opts_chunk$set(fig.align = "center", message = FALSE, warning = FALSE)
options(width = 90)
```

Every calculation from L24 in one runnable place. Last lecture there was one group
and we estimated a single proportion. Today there are two independent groups and
the quantity of interest is the **difference** between their proportions.

There is no external data file. Every number below is typed in from the slides.
Try the practice problems at the bottom before checking the answers.

## The one thing to keep straight

There are two standard errors in this lecture, and they are not interchangeable.

| Purpose | Standard error | Why |
|---------|----------------|-----|
| Confidence interval | $\sqrt{\frac{\hat{p}_1(1-\hat{p}_1)}{n_1} + \frac{\hat{p}_2(1-\hat{p}_2)}{n_2}}$ | no null to lean on, so use the two separate estimates |
| Hypothesis test | $\sqrt{\hat{p}(1-\hat{p})\left(\frac{1}{n_1}+\frac{1}{n_2}\right)}$ | under $H_0$ there is only **one** proportion, so pool |

The pooled estimate is

$$\hat{p} = \frac{\text{successes in both samples}}{\text{individuals in both samples}}$$

Also: **you** choose which group is population 1. Swapping the labels flips the
sign of the difference and reflects the interval about zero. It changes nothing
else. State your choice before you interpret anything.

## Which method when

| Situation | Confidence interval | Test |
|-----------|---------------------|------|
| Successes and failures > 10 in **both** samples | large sample formula | pooled $z$ test |
| Small samples, at least 5 per group | plus four formula | `prop.test()` |
| In R, any time | `prop.test(x, n)` | `prop.test(x, n)` |

# Example 1: the DVT trial (large samples)

Immobilized patients randomized to Fragamin or placebo, outcome is deep vein
thrombosis.

| | DVT | no DVT | Total | $\hat{p}$ |
|---|---|---|---|---|
| Fragamin | 42 | 1476 | 1518 | 0.0277 |
| Placebo | 73 | 1400 | 1473 | 0.0496 |

## Check the condition first

```{r dvt-condition}
c(42, 1476, 73, 1400)      # successes and failures, both arms
```

All four counts are far above 10, so the large sample method applies.

## Large sample confidence interval by hand

Take **population 1 = placebo** and **population 2 = Fragamin**, so a positive
difference means more clots on placebo.

```{r dvt-ci}
p1 <- 73/1473                             # placebo
p2 <- 42/1518                             # Fragamin
n1 <- 1473
n2 <- 1518

se <- sqrt(p1*(1-p1)/n1 + p2*(1-p2)/n2)   # variances add, square root last
c(p1, p2, p1 - p2, se)

(p1 - p2) + c(-1, 1) * 1.96 * se          # 95% CI
```

We are 95% confident the rate of DVT is between 0.8 and 3.6 percentage points
higher on placebo than on Fragamin. The whole interval sits above 0.

Note the pairing under the square root: `n1` goes with `p1`, `n2` with `p2`.
Crossing them is the most common arithmetic error on this formula.

## Hypothesis test by hand

$H_0: p_1 = p_2$ against $H_a: p_1 \neq p_2$.

```{r dvt-test}
p.pool <- (42 + 73) / (1518 + 1473)       # pooled estimate, ALL the data
p.pool

se.0 <- sqrt(p.pool*(1-p.pool)*(1/n1 + 1/n2))   # note: pooled, not p1 and p2
se.0

z <- (p1 - p2) / se.0
z

pnorm(q = z, lower.tail = FALSE) * 2      # two-sided p-value
```

The p-value is about 0.0019. If Fragamin truly did nothing, a gap this large
between the two arms would arise about twice in a thousand trials.

Compare `se` (0.00705) with `se.0` (0.00703). They are close here because the two
proportions are close, but they are computed from genuinely different assumptions.

## The same test in R

```{r dvt-proptest}
prop.test(x = c(42, 73),      # successes
          n = c(1518, 1473))  # sample sizes, same order
```

Read the output bottom up:

- `prop 1` = 0.0277 and `prop 2` = 0.0496 confirm that **Fragamin** went in first
  this time, the opposite of the hand calculation above.
- so the reported CI of -0.036 to -0.007 is the mirror image of the hand interval
  0.007 to 0.036. Same finding, opposite labelling.
- `X-squared` is not a typo. A squared standard Normal is a chi-square with 1
  degree of freedom, and $z^2 = 3.11^2 = 9.69$.

## Matching the hand calculation exactly

R applies a continuity correction by default, which is why its p-value (0.002546)
is slightly larger than ours (0.001853). Turn it off to reproduce the hand answer:

```{r dvt-nocc}
prop.test(x = c(42, 73), n = c(1518, 1473), correct = FALSE)$p.value
prop.test(x = c(42, 73), n = c(1518, 1473), correct = FALSE)$statistic  # = z^2
```

In real work leave the correction on. Turn it off only to check your own
arithmetic.

## Group order changes the sign and nothing else

```{r dvt-order}
prop.test(x = c(42, 73), n = c(1518, 1473))$conf.int   # Fragamin first
prop.test(x = c(73, 42), n = c(1473, 1518))$conf.int   # placebo first
```

Same width, same p-value, mirrored endpoints. This is why a bare number is not an
answer.

# Example 2: the flu vaccine trial (small samples)

| | Flu | no Flu | Total | $\hat{p}$ |
|---|---|---|---|---|
| Vaccine | 4 | 96 | 100 | 0.04 |
| Placebo | 11 | 89 | 100 | 0.11 |

## Why the large sample method is unavailable

```{r flu-condition}
c(4, 96, 11, 89)      # 4 successes is below 10
```

Two hundred people is a respectable trial, but the event is rare. Information here
is carried by the number of **events**, not the number of people.

## Plus four interval by hand

Add 1 success and 1 failure to **each** group, so each denominator gains 2.

```{r flu-plus4}
pt1 <- (4 + 1)/(100 + 2)                  # vaccine
pt2 <- (11 + 1)/(100 + 2)                 # placebo
c(pt1, pt2)

se4 <- sqrt(pt1*(1-pt1)/102 + pt2*(1-pt2)/102)
se4

(pt2 - pt1) + c(-1, 1) * 1.96 * se4       # placebo minus vaccine
```

That is -0.007 to 0.144, or **-0.7% to 14.4%** in percentage points. Watch the
scale. Mixing proportions and percentages is how answers end up off by a factor of
one hundred.

## The same interval from R

```{r flu-proptest}
prop.test(x = c(11, 4), n = c(100, 100))
```

R's interval, -1.2% to 15.2%, is a little wider than the plus four interval
because of the continuity correction. Both contain 0, so both give the same
conclusion.

R never checks the sample size condition. It did not notice that one arm had only
4 events, and it never will. That check is your job.

## Comparing the two intervals

```{r flu-plot, out.width="70%", fig.width=6.5, fig.height=2.8}
flu_CIs <- tibble(
  method   = c("Plus four", "prop.test"),
  lower    = c(-0.7, -1.2),
  upper    = c(14.4, 15.2),
  estimate = c(6.9, 7.0))

ggplot(flu_CIs, aes(x = method, y = estimate)) +
  geom_point(size = 2) +
  geom_segment(aes(xend = method, y = lower, yend = upper)) +
  geom_hline(aes(yintercept = 0), lty = 2) +
  scale_y_continuous(limits = c(-4, 18)) +
  labs(y = "Extra flu cases on placebo\n(percentage points)", x = "") +
  coord_flip() +
  theme_minimal(base_size = 12)
```

Both intervals cross zero, so we cannot rule out that the vaccine does nothing.
But look how much of each interval sits to the right of zero. The honest
conclusion is not "the vaccine does not work". It is "this study is too small to
settle the question".

---

# Practice problems

Try these before looking at the answers.

**Problem 1.** In a cohort study, 150 of 500 exposed people and 120 of 500
unexposed people developed the outcome. (a) Check the condition. (b) Compute the
large sample 95% CI by hand for exposed minus unexposed. (c) Confirm with
`prop.test`. (d) Does the interval contain 0, and what does that imply about a
two-sided test at the 0.05 level?

**Problem 2.** A pilot study finds an adverse event in 2 of 30 patients on drug A
and 8 of 32 on drug B. (a) Why can you not use the large sample interval?
(b) Compute the plus four 95% CI for B minus A. (c) Compare with `prop.test`.

**Problem 3.** 30 of 200 patients on standard care and 45 of 190 on the new
protocol were readmitted. (a) State the hypotheses. (b) Compute the pooled
proportion and the standard error. (c) Compute $z$ and the two-sided p-value by
hand. (d) Confirm with `prop.test(..., correct = FALSE)`. (e) State the
conclusion in the language of the problem.

**Problem 4.** Using the numbers from Problem 3, (a) compute the confidence
interval standard error and the test standard error and say why they differ.
(b) Run `prop.test` with the two groups in each order and describe exactly what
changes.

---

# Answers

## Problem 1

```{r answer-1}
x <- c(150, 120); n <- c(500, 500)

# (a) condition: successes and failures in both groups
c(x, n - x)

# (b) exposed = group 1
p1 <- 150/500; p2 <- 120/500
se <- sqrt(p1*(1-p1)/500 + p2*(1-p2)/500)
c(p1, p2, p1 - p2, se)
(p1 - p2) + c(-1, 1) * 1.96 * se

# (c)
prop.test(x, n)$conf.int
prop.test(x, n)$p.value
```

(a) The four counts are 150, 120, 350 and 380, all far above 10, so the large
sample method is fine.

(b) The difference is 0.06 with a standard error of 0.028, giving a 95% CI of
about 0.5% to 11.5% higher risk in the exposed group.

(d) The interval excludes 0, so a two-sided test at the 0.05 level rejects the
null. The p-value from `prop.test` is about 0.04, which agrees. For a two-sided
test these two statements always travel together.

## Problem 2

```{r answer-2}
# (a) condition
c(2, 30 - 2, 8, 32 - 8)

# (b) plus four, drug B minus drug A
pt.A <- (2 + 1)/(30 + 2)
pt.B <- (8 + 1)/(32 + 2)
se4 <- sqrt(pt.A*(1-pt.A)/32 + pt.B*(1-pt.B)/34)
c(pt.A, pt.B, pt.B - pt.A, se4)
(pt.B - pt.A) + c(-1, 1) * 1.96 * se4

# (c)
prop.test(x = c(8, 2), n = c(32, 30))$conf.int
```

(a) Drug A had only 2 events, well below the 10 required, so the Normal
approximation is not trustworthy. Both groups clear the plus four requirement of
at least 5 observations.

(b) The plus four estimate of the difference is about 17 percentage points, with a
95% CI of roughly -1% to 35%.

(c) `prop.test` gives a similar but slightly wider interval, again because of the
continuity correction. Both contain 0. With 30 patients per arm this study cannot
distinguish a large effect from no effect, which is the usual fate of a pilot.

R also prints `Chi-squared approximation may be incorrect`. That warning is R
noticing the same thing you noticed in part (a), that the counts are too small for
the Normal approximation. It is the one condition check R does volunteer, and it
only appears for the test, never for the interval.

## Problem 3

```{r answer-3}
x <- c(30, 45); n <- c(200, 190)

# (b) pooled proportion and test standard error
p.pool <- sum(x)/sum(n)
se.0 <- sqrt(p.pool*(1-p.pool)*(1/200 + 1/190))
c(p.pool, se.0)

# (c)
p1 <- 30/200; p2 <- 45/190
z <- (p1 - p2)/se.0
c(p1, p2, z)
pnorm(abs(z), lower.tail = FALSE) * 2

# (d)
prop.test(x, n, correct = FALSE)
```

(a) $H_0: p_1 = p_2$, the readmission rate is the same under both protocols,
against $H_a: p_1 \neq p_2$. Two-sided, because a new protocol could plausibly do
harm as well as good.

(b) The pooled proportion is 75/390 = 0.192 and the test standard error is 0.040.

(c) $\hat{p}_1 = 0.15$, $\hat{p}_2 = 0.237$, $z = -2.18$, two-sided p-value about
0.030.

(e) Readmission was about 8.7 percentage points more common under the new
protocol. If the two protocols truly performed the same, a gap this large would
appear about 3 times in 100 studies. That is reasonably strong evidence that the
new protocol readmits more patients, and it is the opposite of what was intended.

## Problem 4

```{r answer-4}
# (a) the two standard errors
se.ci <- sqrt(p1*(1-p1)/200 + p2*(1-p2)/190)     # separate estimates
se.0  <- sqrt(p.pool*(1-p.pool)*(1/200 + 1/190)) # pooled
c(se.ci, se.0)

# (b) order of the groups
prop.test(c(30, 45), c(200, 190), correct = FALSE)$conf.int
prop.test(c(45, 30), c(190, 200), correct = FALSE)$conf.int
prop.test(c(30, 45), c(200, 190), correct = FALSE)$p.value
prop.test(c(45, 30), c(190, 200), correct = FALSE)$p.value
```

(a) They differ because they answer different questions. The confidence interval
has no null hypothesis to lean on, so it estimates each group's variability
separately. The test is computed inside the assumption that the null is true, and
under that assumption there is only one proportion, so all 390 observations
estimate it. Here the two values are 0.03986 and 0.03993, close enough that it
makes no practical difference, but that is a coincidence of this data set and not
a rule.

(b) Only the sign of the confidence interval changes. The endpoints are mirrored
about zero, the width is identical, and the p-value is exactly the same. Which
group you type first is a labelling decision, not a statistical one.
