---
title: "L17: Recap of Part II"
output: html_document
---

```{r load-libraries, include=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
```

This file collects every calculation from L17 in one runnable place. Work the
practice problems at the bottom before you look at the answers in the slides.

## Probability rules

- $0 \leq P(A) \leq 1$
- Probabilities in the sample space sum to 1
- $P(A) + P(\bar{A}) = 1$
- $P(A \cup B) = P(A) + P(B) - P(A \cap B)$
- $P(A) = P(A \cap B) + P(A \cap \bar{B})$
- $P(A \cap B) = P(B|A) \times P(A)$, so $P(B|A) = \frac{P(A \cap B)}{P(A)}$

$A$ and $B$ are independent when $P(A|B) = P(A)$, equivalently
$P(A \cap B) = P(A)P(B)$.

Bayes' theorem:

$$P(A|B) = \frac{P(B|A)P(A)}{P(B|A)P(A) + P(B|A^c)P(A^c)}$$

## Screening tests

Conditional on true disease status, so these describe the test itself:

- Sensitivity = P(Test positive | Disease)
- Specificity = P(Test negative | No disease)

Conditional on the test result, so these move with prevalence:

- Predictive value positive = P(Disease | Test positive)
- Predictive value negative = P(No disease | Test negative)

## Normal distribution

```{r normal-beyonce}
# proportion of US women taller than Beyonce (5'7" = 67 in)
# population: mean 65 in, sd 3.5 in
pnorm(q = 67, mean = 65, sd = 3.5, lower.tail = FALSE)

# same thing via the Z value
z <- (67 - 65) / 3.5
z
pnorm(z, lower.tail = FALSE)

# between Ariana Grande (5'0" = 60 in) and Beyonce (67 in)
pnorm(67, 65, 3.5) - pnorm(60, 65, 3.5)
```

## Binomial distribution

$$P(X = k) = \binom{n}{k} p^k (1-p)^{n-k}$$

```{r binomial-guppies}
# 4 guppies netted, 30% have black spots, want exactly 2 spotted
choose(4, 2)
dbinom(x = 2, size = 4, prob = 0.3)
```

Normal approximation applies when $np \geq 10$ **and** $n(1-p) \geq 10$, with
$X \dot\sim N(\mu = np, \sigma = \sqrt{np(1-p)})$.

## Poisson distribution

$$P(X = k) = \frac{e^{-\mu}\mu^k}{k!}$$

```{r poisson-measles}
# California measles, usual rate 1.75 cases per month
dpois(2, 1.75)                  # exactly 2

dpois(0, 1.75) + dpois(1, 1.75) + dpois(2, 1.75)   # 0, 1 or 2
ppois(2, 1.75)                                     # same thing

# 26 or more. Use lower.tail = FALSE, NOT 1 - ppois(25, 1.75)
ppois(25, 1.75, lower.tail = FALSE)
1 - ppois(25, 1.75)   # returns exactly 0, which is wrong
```

## Sampling distributions

For a sample mean, $\bar{x} \dot\sim N\left(\mu, \frac{\sigma}{\sqrt{n}}\right)$.

For a sample proportion, the standard deviation is $\sqrt{\frac{p(1-p)}{n}}$,
and the Normal approximation needs both $np > 10$ and $n(1-p) > 10$.

## Confidence intervals

$$\bar{x} \pm z^{*}\frac{\sigma}{\sqrt{n}}$$

```{r ci-critical-values}
qnorm(0.95)     # 1.645, for a 90% CI
qnorm(0.975)    # 1.960, for a 95% CI
qnorm(0.995)    # 2.576, for a 99% CI
```

## Power calculation

Storage and sweetness. $\sigma = 1$, $n = 10$, $\alpha = 0.05$, one sided,
and we evaluate power at a true mean change of $-0.8$.

```{r power-sweetness}
# step 1, the cutoff under the null
qnorm(0.05)                              # Z alpha
cutoff <- qnorm(0.05, mean = 0, sd = 1 / sqrt(10))
cutoff

# step 2, the area below that cutoff under the alternative
power <- pnorm(cutoff, mean = -0.8, sd = 1 / sqrt(10))
power
1 - power                                # this is beta
```

```{r power-picture, echo=FALSE, fig.width=7, fig.height=3.6}
se   <- 1 / sqrt(10)
crit <- qnorm(.05, 0, se)
x    <- seq(-2.0, 1.4, length = 1000)

par(mar = c(4, 1, 3, 1))
plot(x, dnorm(x, 0, se), type = "n", ylim = c(0, 1.35),
     ylab = "", xlab = expression(bar(x)), yaxt = "n",
     main = "Null at 0 (solid) vs alternative at -0.8 (dashed)")
xr <- x[x <= crit]
polygon(c(xr, crit, min(x)), c(dnorm(xr, -0.8, se), 0, 0),
        col = "#88AACC", border = NA)
lines(x, dnorm(x,    0, se), lwd = 2)
lines(x, dnorm(x, -0.8, se), lwd = 2, lty = "dashed")
abline(v = crit, lwd = 2, col = "#000088", lty = "dashed")
```

Four things drive sample size: $\alpha$, the effect size, the power you want,
and the variability in the population.

# Practice problems

Seven problems, most of which combine more than one idea. Try each before
looking at the answer slides.

## A. Find the errors

A clinic averages 3.1 medication errors per week. Find P(at least 5 in a week),
and a 95% CI for the mean from 40 weeks with `xbar = 3.1` and `sigma = 1.8`.

```{r practice-a}
# the submitted version had two calculation bugs
1 - ppois(5, lambda = 3.1)          # WRONG, this is P(X >= 6)
1 - ppois(4, lambda = 3.1)          # correct, P(X >= 5)

3.1 + c(-1, 1) * 1.96 * 1.8         # WRONG, used sigma not sigma/sqrt(n)
3.1 + c(-1, 1) * 1.96 * 1.8 / sqrt(40)   # correct
```

The third error was the interpretation. A CI is a statement about the procedure
over repeated samples, not a probability that mu falls in this one interval.

## B. One program, three tools

TB contact tracing. Prevalence 0.06, sensitivity 0.80, specificity 0.97.

```{r practice-b1}
# B1, fraction of positives that are real (PPV)
(0.80 * 0.06) / (0.80 * 0.06 + 0.03 * 0.94)

# B2, false and true positives among 5000 screened
5000 * (1 - 0.06) * (1 - 0.97)
5000 * 0.06 * 0.80

# check that B2 reproduces B1
240 / (240 + 141)
```

```{r practice-b2}
# B3, referrals arrive at 2.4 per hour. P(no referrals in an hour)
dpois(0, lambda = 2.4)

# B4, time to treatment, n = 64, xbar = 21, sigma = 6
21 + c(-1, 1) * qnorm(0.975) * 6 / sqrt(64)
```

## C. What does the output support?

```{r practice-c}
pbinom(3, size = 20, prob = 0.3)                        # P(X <= 3)
dbinom(3, size = 20, prob = 0.3)                        # P(X == 3)
pbinom(3, size = 20, prob = 0.3, lower.tail = FALSE)    # P(X >= 4), NOT P(X >= 3)

1 - pbinom(2, size = 20, prob = 0.3)                    # this is P(X >= 3)
```

## D. Work backwards

```{r practice-d}
# D1, what n gives a 95% CI of total width 5.0 when sigma = 12?
(1.96 * 12 / 2.5)^2        # round UP

# D2, what specificity puts PPV at 0.50 when sens = 0.90 and prevalence = 0.05?
1 - (0.90 * 0.05) / 0.95
```

## E. Rank four studies by power

No computation needed, but the tie is worth verifying. Power depends on the
effect size times the square root of n.

```{r practice-e}
0.4 * sqrt(200)     # Study 3
0.8 * sqrt(50)      # Study 4, identical
```

## F. Six scenarios

Watch scenario 2. The rate is 7 per hour but the window is 15 minutes.

```{r practice-f}
7 / 4                       # lambda for a 15 minute window
dpois(0:6, lambda = 1.75)   # the distribution over that window
```

## G. Answer a colleague

A test is "97% accurate" and the condition affects 1 person in 500.

```{r practice-g}
(0.97 * 0.002) / (0.97 * 0.002 + 0.03 * 0.998)
```

About 6%, not 97%. Reversing a conditional probability changes its value.
